### 2025-05-21 paper.Rmd
converted article draft to Rmd to preview
need to convert back to md file for submission

---
title: 'fairmetrics: Fairness evaluation metrics with confidence intervals'
tags:
  - R
  - Fairness
  - Machine Learning
  - Software
authors:
  - name: Benjamin Smith
    orcid: 0009-0007-2206-0177
    affiliation: 1
  - name: Jianhui Gao
    orcid: 0000-0003-0915-1473
    affiliation: 1
  - name: Benson Chou
    orcid: 0009-0007-0265-033X
    affiliation: 1
  - name: Jessica Gronsbell
    orcid: 0000-0002-5360-5869
    affiliation: 1
affiliations:
  - name: "University of Toronto"
    index: 1
date: "2025-03-08"
bibliography: paper.bib
output:
  rticles::joss_article
  # md_document:
  #   preserve_yaml: TRUE
  #   variant: "markdown_strict"
journal: JOSS
---

# Summary

{fairmetrics} is an R package designed to evaluate the fairness of machine learning models through a range of specialized metrics for which a model can be classified as "fair". It supports fairness assesments of popular group-based criteron, such as independence, seperation, sufficency and others. The package enables statistical inference on fairness metrics through calculation of bootstrap confidence intervals (CIs). In addition, {fairmetrics} offers convenient wrapper functions to compute multiple metrics simultaneously and includes datasets derived from the MIMIC-II clinical database [@goldberger2000physiobank; @raffa2016clinical] for illustrating its use.

# Statement of Need

Machine learning (ML) offers significant potential for predictive modelling in biomedical research [@rajpurkarAIHealthMedicine2022]. Despite its promise, there is substantial evidence that, without appropriate forethought and planning, ML models can introduce or exacerbate health inequities by making less accurate decisions for certain groups or individuals [@grote2022enabling]. As ML becomes increasingly embedded in healthcare systems, ensuring equitable model performance across diverse populations is essential[@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024]. The {fairmetrics} R package allows ML researchers and practitioners to evaluate group fairness of ML models via a suite of popular fairness metrics and provides estimated confidence intervals (CIs) for them through bootstrap estimation.

# Fairness Criteria

A ML can be evaluated as being "fair" through three major criteron: group fairness, individual fairness and causal fairness. Group fairness deems a model fair if its predictions are similarly accurate or callibrated across a predefined set of groups, individual fairness insists that similar individuals should receive similar outcomes, and causal fairness leverages causal models that groups do not have an unjust influence on model predictions [@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024]. {fairmetrics} focuses on calculating group fairness metrics as they are commonly used in biomedical settings. The groups in question are most often defined by protected attributes, such as age or race [@mehrabiSurveyBiasFairness2021].

Group fairness criteria are commonly categorized into three main types: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022]. Independence requires that an ML model's predictions be statistically independent of the protected attribute. Separation demands that the model's predictions be independent of the protected attribute conditional on the true outcome class (i.e., within the positive and negative classes). Sufficiency requires that, given a model's prediction, the likelihood of the true outcome is independent of the protected attribute—aiming to equalize error rates across groups for similar prediction score. The {fairmetrics} package computes a range of group fairness metrics along with bootstrap-based confidence intervals. These metrics are grouped below according to the three core fairness frameworks described above.

## Independence

-   **Statistical Parity:** Compares the overall rate of positive predictions between groups, irrespective of the true outcome.

-   **Conditional Statistical Parity:** Restricts the comparison of positive prediction rates to a specific subgroup (e.g., within a hospital unit or age bracket), offering a more context-specific fairness assessment.

## Separation

-   **Equal Opportunity:** Focuses on disparities in false negative rates (FNR) between two groups, quantifying any difference in missed positive cases.

-   **Predictive Equality:** Compares false positive rates (FPR) between groups, ensuring that no group is disproportionately flagged as positive when the true outcome is negative.

-   **Positive Class Balance:** Checks whether, among individuals whose true outcome is positive, the distribution of predicted probabilities is comparable across groups.

-   **Negative Class Balance:** Checks whether, among individuals whose true outcome is negative, the distribution of predicted probabilities is comparable across groups.

## Sufficiency

-   **Predictive Parity:** Compares positive predictive values (PPV) across groups, assessing whether the precision of positive predictions is equivalent.

## Other Criteria

-   **Brier Score Parity:** Assesses whether the Brier score—the mean squared error of probabilistic predictions—is similar across groups, indicating comparable calibration.

-   **Accuracy Parity:** Measures whether the overall accuracy of a predictive model is equivalent across different groups.

-   **Treatment Equality:** Compares the ratio of false negatives to false positives across groups, ensuring the balance of missed detections versus false alarms is consistent.

# Main Features

This

# Additional Features

Beyond individual metric computation, the {fairmetrics} package includes convenience functions to retrieve multiple fairness metrics (and their confidence intervals) in a single call. It also bundles example datasets based on the the MIMIC-II clinical database [@goldberger2000physiobank; @raffa2016clinical] to facilitate simple example usage of the fairness metrics with ML models.

# Related Work

Other R packages similar to {fairmetrics} include {fairness}[@fairness_package] and {fairmodels}[@wisniewski2022fairmodels]. The differences between {fairmetrics} and these other packages is twofold. The primary difference between is that {fairmetrics} allow for the calculation of estimated confidence intervals of fairness metrics via bootstrap, which allows for more meaningful inferences about the fairness metrics calculated. Additionally, the {fairness} package has fewer dependencies and a lower memory footprint, making the for a more environment agnostic tool that can be used with modest hardware.

<!--
Jesse: Should I add the table which I genereated in README here?
--->

# Licensing and Availability

The {fairmetrics} package is under the MIT liscence[@mit_license] and is available on CRAN and Github. The CRAN release can be installed with `install.packages("fairmetrics")`. For installing from Github, the {devtools} package[@devtools_package] or any other R package which allows for installation of packages hosted on Github can be used (i.e.`devtools::install_github("jianhuig/fairmetrics")`).

# References

### 2025-05-22 paper.Rmd
adding table and refs for figures/tables

@@
 ---
-title: 'fairmetrics: Fairness evaluation metrics with confidence intervals'
+title: '{fairmetrics}: An R package for fairness evaluation metrics with confidence intervals'
 tags:
@@
 {fairmetrics} is an R package designed to evaluate the fairness of machine learning models through a range of specialized metrics for which a model can be classified as "fair".
-It supports fairness assesments of popular group-based criteron, such as independence, seperation, sufficency and others.
+It supports fairness assessments of popular group-based criteron, such as independence, separation, sufficency and others.
 The package enables statistical inference on fairness metrics through calculation of bootstrap confidence intervals (CIs).
@@
 Despite its promise, there is substantial evidence that, without appropriate forethought and planning, ML models can introduce or exacerbate health inequities by making less accurate decisions for certain groups or individuals [@grote2022enabling].
+While existing software can compute fairness metrics, none provide out-of-the-box statistical inference, leaving practitioners without guidance on the uncertainty around those metrics.
 As ML becomes increasingly embedded in healthcare systems, ensuring equitable model performance across diverse populations is essential[@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
-The {fairmetrics} R package allows ML researchers and practitioners to evaluate group fairness of ML models via a suite of popular fairness metrics and provides estimated confidence intervals (CIs) for them through bootstrap estimation.
+The {fairmetrics} R package fills this gap by offering a suite of popular group-fairness metrics along with bootstrap-based confidence intervals, enabling more rigorous and interpretable assessments of fairness in biomedical ML.
 
@@
 
-A ML can be evaluated as being "fair" through three major criteron: group fairness, individual fairness and causal fairness.
+An ML model can be evaluated as being "fair" through three major criteron: group fairness, individual fairness and causal fairness.
 Group fairness deems a model fair if its predictions are similarly accurate or callibrated across a predefined set of groups, individual fairness insists that similar individuals should receive similar outcomes, and causal fairness leverages causal models that groups do not have an unjust influence on model predictions  [@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
@@
 
-# Main Features
+# Evaluating Fairness Criteria
 
-This
+The {fairmetrics} package requires that a model has been trained and validated with data that has been split appropriately.
+The primary required input is a data frame or tibble which containing the model predictions, true outcomes, and protected attributes.
+\hyperref[workflow]{Figure ~\ref*{workflow}}
+ shows the workflow for using {fairmetrics}. It is possible to evaluate a model for a specific or multiple group fairness metrics.
 
-# Additional Features
+![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow.png)
 
-Beyond individual metric computation, the {fairmetrics} package includes convenience functions to retrieve multiple fairness metrics (and their confidence intervals) in a single call.
-It also bundles example datasets based on the the MIMIC-II clinical database [@goldberger2000physiobank; @raffa2016clinical] to facilitate simple example usage of the fairness metrics with ML models.
+More concretely, an example of how to use the {fairmetrics} package is shown below.
+The example uses the `mimic_preprocessed` dataset, which is a preprocessed version of the MIMIC-II database [@goldberger2000physiobank; @raffa2016clinical].
+
+Consider a random forest model trained to predict the risk of heart failure in patients.
+Suppose we are interested in evaluating the model's fairness with respect to patient gender with predictive equality.
+
+```r
+library(fairmetrics)
+library(dplyr)
+library(magrittr)
+library(randomForest)
+
+# Load the example dataset
+data("mimic_preprocessed")
+
+# Split the data into training and test sets
+train_data <- mimic_preprocessed %>%
+  dplyr::filter(dplyr::row_number() <= 700)
+
+test_data <- mimic_preprocessed %>%
+  dplyr::mutate(gender = ifelse(gender_num == 1, "Male", "Female")) %>%
+  dplyr::filter(dplyr::row_number() > 700)
+
+# Train a random forest model
+rf_model <- randomForest::randomForest(
+  factor(day_28_flg) ~ ., 
+  data = train_data, 
+  ntree = 1000
+  )
+
+# Make predictions on the test set
+test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")
+
+# Evaluate predictive equality
+# (Setting message=FALSE to avoid cluttering the output)
+
+eval_pred_equality(
+  data = test_data,
+  outcome = "day_28_flg",
+  group = "gender",
+  probs = "pred",
+  cutoff = 0.41,
+  message = FALSE
+)
+
+#>   Metric GroupFemale GroupMale Difference  95% Diff CI Ratio 95% Ratio CI
+#> 1    FPR        0.08      0.03       0.05 [0.02, 0.08]  2.67 [1.38, 5.15]
+
+```
+This case it can be seen from both the difference and ratio of the false positive rates (FPR) that the model is biased, with a 2.67 times higher chance of a false positive prediction for heart failure for females compared to males.
+The 95% bootstrap confidence intervals for both the difference and the ratio of the FPR are also shown, which can be used to further assess the statistical significance of the difference, confirming the initial evaluation of the FPR for female and male patients.
 
@@
 The differences between {fairmetrics} and these other packages is twofold.
-The primary difference between is that {fairmetrics} allow for the calculation of estimated confidence intervals of fairness metrics via bootstrap, which allows for more meaningful inferences about the fairness metrics calculated.
+The primary difference between is that {fairmetrics} allows for the calculation of estimated confidence intervals of fairness metrics via bootstrap, which allows for more meaningful inferences about the fairness metrics calculated.
 Additionally, the {fairness} package has fewer dependencies and a lower memory footprint, making the for a more environment agnostic tool that can be used with modest hardware.
+\hyperref[tab:memory_usage]{Table~\ref*{tab:memory_usage}} shows the memory comparison when loading each library.
 
-<!--
-Jesse: Should I add the table which I genereated in README here?
---->
+\begin{table}[ht]
+\centering
+\begin{tabular}{l r}
+\hline
+\textbf{Package} & \textbf{Memory (MB)} \\
+\hline
+fairmodels  & 17.02  \\
+fairness    & 117.61 \\
+fairmetrics & 0.05   \\
+\hline
+\end{tabular}
+\caption{Memory usage of {fairmetrics} vs similar packages (MB)}
+\label{tab:memory_usage}
+\end{table}
+
+For python users, the {fairlearn} library [@fairlearn_paper] provides a broader set of fairness metrics and algorithms.
+The {fairmetrics} package is designed for seemless integration with R workflows, making it a more convenient choice for R-based ML applications.
 
@@
 
-The {fairmetrics} package is under the MIT liscence[@mit_license] and is available on CRAN and Github.
-The CRAN release can be installed with `install.packages("fairmetrics")`.
-For installing from Github, the {devtools} package[@devtools_package] or any other R package which allows for installation of packages hosted on Github can be used (i.e.`devtools::install_github("jianhuig/fairmetrics")`).
+The {fairmetrics} package is under the MIT license.
+It is available on CRAN and can be installed by using `install.packages("fairmetrics")`.
+A more in-depth tutorial can be accessed at: https://jianhuig.github.io/fairmetrics/articles/fairmetrics.html.
+All code is open-source and hosted on
+GitHub. All bugs and inquiries can be reported at https://github.com/jianhuig/fairmetrics/issues/.
 

### 2025-05-28 paper.Rmd
adding jesse's comments from the md file to the rmd file.
Using HTML comments for Rmd

@@
 
-{fairmetrics} is an R package designed to evaluate the fairness of machine learning models through a range of specialized metrics for which a model can be classified as "fair".
-It supports fairness assessments of popular group-based criteron, such as independence, separation, sufficency and others.
-The package enables statistical inference on fairness metrics through calculation of bootstrap confidence intervals (CIs).
-In addition, {fairmetrics} offers convenient wrapper functions to compute multiple metrics simultaneously and includes datasets derived from the MIMIC-II clinical database [@goldberger2000physiobank; @raffa2016clinical] for illustrating its use.
+<!--- 
+Rewritten by Jesse
+--->
+Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for certain groups, particularly those defined by protected attributes such as race, gender, or age.
+Evaluating fairness is a critical aspect of model development, as biased models can perpetuate or exacerbate existing social inequalities.
+The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity).
+These criteria assess whether a model is equally accurate or well-calibrated across predefined groups so that appropriate bias mitigation strategies can be implemented.
+{fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example an dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
 
 # Statement of Need
+
+<!---
+ianhui - this is too similar to the original paper still and I suggest we emphasize that the package isn't restricted to biomedical data.
+We can of course use it as an example but this isn't the need for the package.
+Can you rewrite this to just be about the general use of fairness?
+i.e., point 1 - ml increasing integrated into different areas, point 2 - give a few examples, point 3 - need for the package in light of other work.
+I don't think interpetable is the correct word here, the key difference is allowing for uncertainty quantification, we also provide ratio + difference estimates, i dont think R fairness package does, I would emphasize
+--->
 
@@
 
-An ML model can be evaluated as being "fair" through three major criteron: group fairness, individual fairness and causal fairness.
+<!---
+# REMOVED: not relevant to the package
+An ML model can be evaluated as being "fair" through three major criterion: group fairness, individual fairness and causal fairness.
 Group fairness deems a model fair if its predictions are similarly accurate or callibrated across a predefined set of groups, individual fairness insists that similar individuals should receive similar outcomes, and causal fairness leverages causal models that groups do not have an unjust influence on model predictions  [@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
@@
 The groups in question are most often defined by protected attributes, such as age or race [@mehrabiSurveyBiasFairness2021].
+--->
 
-Group fairness criteria are commonly categorized into three main types: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022].
-Independence requires that an ML model's predictions be statistically independent of the protected attribute.
-Separation demands that the model's predictions be independent of the protected attribute conditional on the true outcome class (i.e., within the positive and negative classes).
-Sufficiency requires that, given a model's prediction, the likelihood of the true outcome is independent of the protected attribute—aiming to equalize error rates across groups for similar prediction score.
-The {fairmetrics} package computes a range of group fairness metrics along with bootstrap-based confidence intervals.
-These metrics are grouped below according to the three core fairness frameworks described above.
+Group fairness criteria are typically classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022].
+The {fairmetrics} package computes a range of group fairness metrics together with bootstrap-based confidence intervals for uncertainty quantification.
+The metrics implemented in the package are briefly described below.
+
+<!---
+1.
+Jianhui - The scope of the package is not defined in the above paragprah so one would not understand the definitons below (i.e., "positive classification" is never defined).
+Specifically, you need to say you consider binary classification and a binary protected attribute - I do not believe your package handles more than.
+Explain why this is done and then update the definitions below to reflect this.
+
+2. Jianhui - why do you use bootstrap for this package rather than IF?  Is it because you don't have IF for all metrics?
+
+3. Jianhui - please update the initial sentences describing the 3 categories to be more intuitive, akin to what we have in the paper.  
+--->
 
@@
 
+<!---
+Use consistent language throughout - alternating between checks/compares/focuses/assess/measure/etc is confusing as all functions do the same thing.  Suggest use "Compares" for all. 
+--->
+
 ## Separation
@@
 
+<!---
+## Do you need the acronyms for FNR and FPR? If you don't use them later, they don't need to be defined.
+--->
 -   **Predictive Equality:** Compares false positive rates (FPR) between groups, ensuring that no group is disproportionately flagged as positive when the true outcome is negative.
@@
 
-The {fairmetrics} package requires that a model has been trained and validated with data that has been split appropriately.
-The primary required input is a data frame or tibble which containing the model predictions, true outcomes, and protected attributes.
-\hyperref[workflow]{Figure ~\ref*{workflow}}
- shows the workflow for using {fairmetrics}. It is possible to evaluate a model for a specific or multiple group fairness metrics.
+The input to the {fairmetrics} package is a data frame or tibble which containing the model's predictions, true outcomes, and the protected attributes (attribute?
+doesn't package only handle 1?).
+\hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}.
+It is possible to evaluate a model for a specific or multiple group fairness metrics.
 
@@
 
-More concretely, an example of how to use the {fairmetrics} package is shown below.
-The example uses the `mimic_preprocessed` dataset, which is a preprocessed version of the MIMIC-II database [@goldberger2000physiobank; @raffa2016clinical].
+<!---
+Edited by Jesse
+--->
+A simple example of how to use the {fairmetrics} package is shown below.
+The example uses the `mimic_preprocessed` dataset, which is a pre-processed version of the MIMIC-II database [@goldberger2000physiobank; @raffa2016clinical].
 
-Consider a random forest model trained to predict the risk of heart failure in patients.
-Suppose we are interested in evaluating the model's fairness with respect to patient gender with predictive equality.
+<!---
+1. Ben - need to add detail on what the detail is in the above paragraph. What does the data contaon, where did it come from, what is the sample size, etc.
+
+2.
+Ben - suggest you show more of the functions in the package - say that, while the choice of metric is context dependent, we show all metrics to show the full range of the package.
+Also include interpretation of the various results.
+Simple way to do this is show, the multiple output then show an example of one of the separation-based criterion or other to show users they can specify what they want.
+---->
 
@@
 
-Other R packages similar to {fairmetrics} include {fairness}[@fairness_package] and {fairmodels}[@wisniewski2022fairmodels].
+<!---
+## Both - you are missing this one: https://mlr3fairness.mlr-org.com
+## Do these metrics give difference and ratio based criteria? Are there any other differences? 
+--->
+Other R packages similar to {fairmetrics} include {fairness}[@fairness_package] and {fairmodels} [@wisniewski2022fairmodels].
 The differences between {fairmetrics} and these other packages is twofold.
 The primary difference between is that {fairmetrics} allows for the calculation of estimated confidence intervals of fairness metrics via bootstrap, which allows for more meaningful inferences about the fairness metrics calculated.
-Additionally, the {fairness} package has fewer dependencies and a lower memory footprint, making the for a more environment agnostic tool that can be used with modest hardware.
-\hyperref[tab:memory_usage]{Table~\ref*{tab:memory_usage}} shows the memory comparison when loading each library.
+Additionally, in contrast to the {fairmodels} and {fairness} packages, the {fairmetrics} package has zero library dependencies and a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems.
+\hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows the comparison of memory used and dependencies required when loading each library.
 
@@
 \centering
-\begin{tabular}{l r}
+\begin{tabular}{l r r}
 \hline
-\textbf{Package} & \textbf{Memory (MB)} \\
+\textbf{Package} & \textbf{Memory (MB)} & \textbf{Dependencies} \\
 \hline
-fairmodels  & 17.02  \\
-fairness    & 117.61 \\
-fairmetrics & 0.05   \\
+fairmodels  & 17.02  & 29\\
+fairness    & 117.61 & 141\\
+fairmetrics & 0.05   & 0\\
 \hline
 \end{tabular}
-\caption{Memory usage of {fairmetrics} vs similar packages (MB)}
-\label{tab:memory_usage}
+\caption{Memory usage and dependencies of {fairmetrics} vs similar packages (MB)}
+\label{tab:memory_dep_usage}
 \end{table}

### 2025-05-28 R/get_fairness_metrics.R
noticed typo in get_fairness_metrics

@@ -58,7 +58,7 @@
 #' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
-#' # Evaluate Accuracy Parity
+#' # Get Fairness Metrics
 #' get_fairness_metrics(
 #'  data = test_data,
 #'  outcome = "day_28_flg",

### 2025-05-28 paper.Rmd
adding edits
still more to go

@@
 <!---
-ianhui - this is too similar to the original paper still and I suggest we emphasize that the package isn't restricted to biomedical data.
+Jianhui - this is too similar to the original paper still and I suggest we emphasize that the package isn't restricted to biomedical data.
 We can of course use it as an example but this isn't the need for the package.
@@
 <!---
-# REMOVED: not relevant to the package
-An ML model can be evaluated as being "fair" through three major criterion: group fairness, individual fairness and causal fairness.
-Group fairness deems a model fair if its predictions are similarly accurate or callibrated across a predefined set of groups, individual fairness insists that similar individuals should receive similar outcomes, and causal fairness leverages causal models that groups do not have an unjust influence on model predictions  [@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
-{fairmetrics} focuses on calculating group fairness metrics as they are commonly used in biomedical settings.
-The groups in question are most often defined by protected attributes, such as age or race [@mehrabiSurveyBiasFairness2021].
+This section needs to be cleaned up for clarity - see comments below.
 --->
@@
 Use consistent language throughout - alternating between checks/compares/focuses/assess/measure/etc is confusing as all functions do the same thing.  Suggest use "Compares" for all.
+
+Ben Response: Rewritten to use "Compares".
 --->
@@
 
--   **Equal Opportunity:** Focuses on disparities in false negative rates (FNR) between two groups, quantifying any difference in missed positive cases.
+-   **Equal Opportunity:**  Compares disparities in false negative rates between two groups, quantifying any difference in missed positive cases.
 
 <!---
-## Do you need the acronyms for FNR and FPR? If you don't use them later, they don't need to be defined.
+Do you need the acronyms for FNR and FPR? If you don't use them later, they don't need to be defined.
+
+Ben Response: FPR is used. FNR is not. Removed FNR. 
 --->
@@
 
--   **Positive Class Balance:** Checks whether, among individuals whose true outcome is positive, the distribution of predicted probabilities is comparable across groups.
+-   **Positive Class Balance:**  Compares the distribution of predicted probabilities among individuals whose true outcome is positive between groups, ensuring that the model does not favor one group over another in its positive predictions.
 
--   **Negative Class Balance:** Checks whether, among individuals whose true outcome is negative, the distribution of predicted probabilities is comparable across groups.
+-   **Negative Class Balance:** Compares the distribution of predicted probabilities among individuals whose true outcome is negative between groups, ensuring that the model does not favor one group over another in its negative predictions.
 
@@
 
--   **Predictive Parity:** Compares positive predictive values (PPV) across groups, assessing whether the precision of positive predictions is equivalent.
+-   **Predictive Parity:** Compares positive predictive values across groups, assessing whether the precision of positive predictions is equivalent.
 
@@
 
--   **Brier Score Parity:** Assesses whether the Brier score—the mean squared error of probabilistic predictions—is similar across groups, indicating comparable calibration.
+-   **Brier Score Parity:** Compares the Brier score—the mean squared error of probabilistic predictions—is similar across groups, indicating comparable calibration.
 
--   **Accuracy Parity:** Measures whether the overall accuracy of a predictive model is equivalent across different groups.
+-   **Accuracy Parity:** Compares the overall accuracy of a predictive model is equivalent across different groups.
 
@@
 
-The input to the {fairmetrics} package is a data frame or tibble which containing the model's predictions, true outcomes, and the protected attributes (attribute?
-doesn't package only handle 1?).
+<!--
+1.
+Ben - The package doesn't require data splitting, some people may want to use the package and incorporate CV within training.
+Also, sometimes people evaluate fairness metrics on validation data.
+So suggest you simplify text + figure.
+Just start with the input.
+Also, it doesn't seem like multiple should be a separate node?
+doesnt it just include all the 4 boxes (sep, suff, ind, other)?
+can this be depicted?
+Also suggest you make the names more descriptive "Separation-based Metrics", "Independence-based Metrics", etc.
+--->
+The primary input to the {fairmetrics} package is a data frame or tibble which containing the model's predictions, true outcomes, and the protected attribute in question.
 \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}.
@@
 A simple example of how to use the {fairmetrics} package is shown below.
-The example uses the `mimic_preprocessed` dataset, which is a pre-processed version of the MIMIC-II database [@goldberger2000physiobank; @raffa2016clinical].
+The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data].
+This dataset consists of 1776 hemodynamically stable patients with resperatory failure, and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating wheter the patient died within 28 days of admission.
 
+While the choice of fairness metric used is context dependent, we show all metrics available with the `get_fairness_metrics()` function.
+In this example, we evaluate the model's fairness with respect to the protected attribute `gender`.
+For conditional statistical parity, we condition  being `>=60` years old.
+The model is trained on a subset of the data, and predictions are made on a test set.
 <!---
 1. Ben - need to add detail on what the detail is in the above paragraph. What does the data contaon, where did it come from, what is the sample size, etc.
+
+Ben Response: Done.
 
@@
 Simple way to do this is show, the multiple output then show an example of one of the separation-based criterion or other to show users they can specify what they want.
+Ben Response: Please see below for the example and review.
 ---->
@@
 
-eval_pred_equality(
+get_fairness_metrics(
+ data = test_data,
+ outcome = "day_28_flg",
+ group = "gender",
+ group2 = "age",
+ condition = ">=60",
+ probs = "pred",
+ cutoff = 0.41
+)
+
+#>                                       Metric GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
+#> 1                         Statistical Parity        0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
+#> 2  Conditional Statistical Parity (age >=60)        0.34      0.21       0.13   [0.05, 0.21]  1.62 [1.18, 2.22]
+#> 3                          Equal Opportunity        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
+#> 4                        Predictive Equality        0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.39, 5.1]
+#> 5                 Balance for Positive Class        0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.41]
+#> 6                 Balance for Negative Class        0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
+#> 7                          Predictive Parity        0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.23]
+#> 8                         Brier Score Parity        0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
+#> 9                    Overall Accuracy Parity        0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
+#> 10                        Treatment Equality        1.03      3.24      -2.21 [-4.35, -0.07]  0.32 [0.15, 0.68]
+```
+
+Among the fairness metrics calculated, metrics whose difference confidence intervals cross zero and whose ratio confidence intervals cross one indicate no significant difference between the groups.
+In this example, the statistical parity, conditional statistical pairty, equal opportunity, predictive equality, positive class balance, negative class balance, and treatment equaltiy metrics show a significant differences.
+While, the predictive parity, Brier score parity, and overall accuracy parity metrics do not show significant differences between the groups.
+
+Should the user wish to calculate an individual metric, it is possible to use any of the `eval_*` functions.
+For example, to calculate the equal opportunity metric, the user can use the `eval_equal_opportunity()` function.
+
+```r
+eval_eq_opp(
   data = test_data,
@@
   probs = "pred",
-  cutoff = 0.41,
-  message = FALSE
+  cutoff = 0.41
 )
 
+#>There is evidence that model does not satisfy equal opportunity.
 #>  Metric GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#> 1    FPR        0.08      0.03       0.05 [0.02, 0.08]  2.67 [1.38, 5.15]
+#>1    FNR        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.85]
+```
 
+For more traditional fairness metrics, such as the True Positive Rate (TPR), False Positive Rate (FPR), Positive Predictive Value (PPV), Negative Predictive Value (NPV), and others, the `get_all_metrics()` function can be used.
+This function returns a data frame with the metrics calculated for each group.
+
+```r
+get_all_metrics(
+  dat = test_data,
+  outcome = "day_28_flg",
+  group = "gender",
+  probs = "pred",
+  cutoff = 0.41
+)
+
+#>         Metric Group Female Group Male
+#> 1           TPR         0.62       0.38
+#> 2           FPR         0.08       0.03
+#> 3           PPR         0.17       0.08
+#> 4           PPV         0.62       0.66
+#> 5           NPV         0.92       0.90
+#> 6           ACC         0.87       0.88
+#> 7   Brier Score         0.09       0.08
+#> 8   FN/FP Ratio         1.03       3.24
+#> 9 Avg Pred Prob         0.21       0.14
 ```
-This case it can be seen from both the difference and ratio of the false positive rates (FPR) that the model is biased, with a 2.67 times higher chance of a false positive prediction for heart failure for females compared to males.
-The 95% bootstrap confidence intervals for both the difference and the ratio of the FPR are also shown, which can be used to further assess the statistical significance of the difference, confirming the initial evaluation of the FPR for female and male patients.
 
@@
 ## Do these metrics give difference and ratio based criteria? Are there any other differences?
+
+Ben Response: Added mlr3fairness package updated the differences.
+
+For comparison:
+
+- {fairness}- does not give difference or ratio based criteria, only point estimates. The package also produces plots which {fairmetrics} does not (not in our scope).
+- {fairmodels}- does give ratio criteria but not GROUP ratio based criteria, only overall ratio.
+The package also produces plots which {fairmetrics} does not (not in our scope).
+No difference based criteria.
+- Same with {mlr3fairness} - does not give difference or ratio based criteria, only point estimates. The package also produces plots which {fairmetrics} does not (not in our scope).
+
+(I would also add that the {fairmodels} has added complexity required to using it, and {mlr3fairness} packages are designed for use with the mlr3 framework, which is not as widely adopted in the R community.
+This could be a "skill issue" on my part.
+However I don't think such a comment would be appropriate for the paper.)
+
 --->
-Other R packages similar to {fairmetrics} include {fairness}[@fairness_package] and {fairmodels} [@wisniewski2022fairmodels].
+Other R packages similar to {fairmetrics} include {fairness}[@fairness_package], {fairmodels} [@wisniewski2022fairmodels] and {mlr3fairness}[mlr3fairness_package].
 The differences between {fairmetrics} and these other packages is twofold.
-The primary difference between is that {fairmetrics} allows for the calculation of estimated confidence intervals of fairness metrics via bootstrap, which allows for more meaningful inferences about the fairness metrics calculated.
-Additionally, in contrast to the {fairmodels} and {fairness} packages, the {fairmetrics} package has zero library dependencies and a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems.
+The primary difference between is that {fairmetrics} calculates the ratio and difference between group fairness criterion and allows estimated confidence intervals of fairness metrics via bootstrap - allowing for more meaningful inferences about the fairness metrics calculated.
+Additionally, in contrast to the {fairmodels}, {fairness} and {mlr3fairness} packages, the {fairmetrics} package does not posses any external dependencies and has a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems.
 \hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows the comparison of memory used and dependencies required when loading each library.
@@
 \hline
-fairmodels  & 17.02  & 29\\
+fairmodels  & 17.02  & 29 \\
 fairness    & 117.61 & 141\\
-fairmetrics & 0.05   & 0\\
+fairmodels  & 58.11  & 45 \\
+fairmetrics & 0.05   & 0  \\
 \hline

### 2025-05-29 paper.Rmd
Update paper.Rmd

@@
 These criteria assess whether a model is equally accurate or well-calibrated across predefined groups so that appropriate bias mitigation strategies can be implemented.
-{fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example an dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
+{fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example a dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
 
@@
 
-Machine learning (ML) offers significant potential for predictive modelling in biomedical research [@rajpurkarAIHealthMedicine2022].
-Despite its promise, there is substantial evidence that, without appropriate forethought and planning, ML models can introduce or exacerbate health inequities by making less accurate decisions for certain groups or individuals [@grote2022enabling].
-While existing software can compute fairness metrics, none provide out-of-the-box statistical inference, leaving practitioners without guidance on the uncertainty around those metrics.
-As ML becomes increasingly embedded in healthcare systems, ensuring equitable model performance across diverse populations is essential[@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
-The {fairmetrics} R package fills this gap by offering a suite of popular group-fairness metrics along with bootstrap-based confidence intervals, enabling more rigorous and interpretable assessments of fairness in biomedical ML.
+Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education.
+As ML systems become a commonplace in decision-making processes, mounting evidence suggest that they often exhibit bias towards gender and ethnicity.
+For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants.
+As COMPAS was used in court to inform judicial decision-making, this bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to White defendants with similar risk profiles.
+Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients.
+This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to White patients.
+Therefore, there is a urgent need for companies, engineers, and researchers to ensure that ML models are fair and equitable across different groups before they are deployed in real-world applications.
+
+While existing software can compute fairness metrics, they provide only point estimates without quantifying the uncertainty around those metrics.
+This limitation prevents users from determining whether observed disparities between groups are statistically significant or merely the result of random variation due to finite sample size, potentially leading to incorrect conclusions about fairness violations.
+The {fairmetrics} R package addresses this gap by providing bootstrap-based confidence intervals (CIs) for both difference-based and ratio-based group fairness metrics, enabling users to make statistically grounded decisions about the fairness of their models.
+
+# Scope
+
+The `{fairmetrics}` package is designed to evaluate group fairness in the setting of binary classification with a binary protected attribute.
+This restriction reflects standard practice in the fairness literature and is motivated by several considerations.
+First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative.
+Second, group fairness is the most widely used framework for binary classification tasks.
+Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness.
+This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.
 
@@
 Group fairness criteria are typically classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022].
-The {fairmetrics} package computes a range of group fairness metrics together with bootstrap-based confidence intervals for uncertainty quantification.
-The metrics implemented in the package are briefly described below.
+Independence requires that the model’s predictions be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.
+Separation requires independence between the prediction and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within positive (or negative) outcome class.
+Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model’s prediction is known, the protected attribute provides no additional information about the true outcome.
 
@@
 
--   **Equal Opportunity:**  Compares disparities in false negative rates between two groups, quantifying any difference in missed positive cases.
+-   **Equal Opportunity:**  Compares disparities in false negative rates between groups, quantifying any difference in missed positive cases.
 
@@
 
--   **Positive Class Balance:**  Compares the distribution of predicted probabilities among individuals whose true outcome is positive between groups, ensuring that the model does not favor one group over another in its positive predictions.
+<!---
+Updated the two defs below. It only checks the first-moment of the predicted probabilities, does not check the distribution.
 
--   **Negative Class Balance:** Compares the distribution of predicted probabilities among individuals whose true outcome is negative between groups, ensuring that the model does not favor one group over another in its negative predictions.
+Also updated the name, Balance of Positive/Negative Class is the name the original paper used.
+--->
+
+-   **Balance for Positive Class:**  Compares if the average of predicted probabilities among individuals whose true outcome is positive is similar between groups.
+
+-   **Balance for Negative Class:** Compares if the average of predicted probabilities among individuals whose true outcome is negative is similar between groups.
 
@@
 
--   **Brier Score Parity:** Compares the Brier score—the mean squared error of probabilistic predictions—is similar across groups, indicating comparable calibration.
+-   **Brier Score Parity:** Compares the Brier score—the mean squared error of predicted probabilities—is similar across groups, indicating comparable calibration.
 
--   **Accuracy Parity:** Compares the overall accuracy of a predictive model is equivalent across different groups.
+-   **Accuracy Parity:** Compares the overall accuracy of a predictive model is equivalent across groups.
 
@@
 Also suggest you make the names more descriptive "Separation-based Metrics", "Independence-based Metrics", etc.
+
+Ben Response: Created 3 different figures to review. I think v3 is the best. Please critique if there are any issues.
 --->
@@
 
-![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow.png)
+![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow-v3.png)
 
 <!---
-Edited by Jesse
+Previously edited by Jesse, now edited by Ben
 --->
@@
 The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data].
-This dataset consists of 1776 hemodynamically stable patients with resperatory failure, and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating wheter the patient died within 28 days of admission.
+This dataset consists of 1776 hemodynamically stable patients with respiratory failure, and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating whether the patient died within 28 days of admission.
 
@@
 In this example, we evaluate the model's fairness with respect to the protected attribute `gender`.
-For conditional statistical parity, we condition  being `>=60` years old.
+For conditional statistical parity, we condition on patients older than 60 years old.
 The model is trained on a subset of the data, and predictions are made on a test set.
@@
 Simple way to do this is show, the multiple output then show an example of one of the separation-based criterion or other to show users they can specify what they want.
-Ben Response: Please see below for the example and review.
+
+Ben Response: Please see below for the examples and review.
 ---->

### 2025-05-29 paper.Rmd
Update paper.Rmd

@@
 
-For more traditional fairness metrics, such as the True Positive Rate (TPR), False Positive Rate (FPR), Positive Predictive Value (PPV), Negative Predictive Value (NPV), and others, the `get_all_metrics()` function can be used.
+For group-specific performance metrics, such as the True Positive Rate (TPR), False Positive Rate (FPR), Positive Predictive Value (PPV), Negative Predictive Value (NPV), and others, the `get_all_metrics()` function can be used.
 This function returns a data frame with the metrics calculated for each group.

### 2025-05-29 paper.Rmd
Update paper.Rmd

@@
 affiliations:
-  - name: "University of Toronto"
+  - name: "Department of Statistcial Science, University of Toronto"
     index: 1

### 2025-05-29 paper.Rmd
Update paper.Rmd

--- 2025-05-29 paper.Rmd
@@
     affiliation: 1
-  - name: Benson Chou
-    orcid: 0009-0007-0265-033X
-    affiliation: 1
   - name: Jessica Gronsbell

### 2025-05-29 paper.Rmd
add citations

@@
 
-Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education.
+Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21].
 As ML systems become a commonplace in decision-making processes, mounting evidence suggest that they often exhibit bias towards gender and ethnicity.
-For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants.
+For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants[@mattuMachineBias].
 As COMPAS was used in court to inform judicial decision-making, this bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to White defendants with similar risk profiles.
-Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients.
+Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients[@obermeyerDissectingRacialBias2019].
 This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to White patients.
@@
 This restriction reflects standard practice in the fairness literature and is motivated by several considerations.
-First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative.
-Second, group fairness is the most widely used framework for binary classification tasks.
+First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative[@mehrabi_survey_21].
+Second, group fairness is the most widely used framework for binary classification tasks[@mehrabi_survey_21].
 Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness.

### 2025-05-30 paper.Rmd
Update title + summary.

@@
 ---
-title: '{fairmetrics}: An R package for fairness evaluation metrics with confidence intervals'
+title: '{fairmetrics}: An R package for group fairness evaluation'
 tags:

### 2025-05-30 paper.Rmd
Edit Statement of Need.

@@
 
-<!--- 
-Rewritten by Jesse
---->
-Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for certain groups, particularly those defined by protected attributes such as race, gender, or age.
-Evaluating fairness is a critical aspect of model development, as biased models can perpetuate or exacerbate existing social inequalities.
+Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for particular groups, particularly those defined by protected attributes such as race, gender, or age.
+Evaluating fairness is a critical aspect of ML model development, as biased models can perpetuate structural inequalities.
 The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity).
-These criteria assess whether a model is equally accurate or well-calibrated across predefined groups so that appropriate bias mitigation strategies can be implemented.
-{fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example a dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
+Group-based fairness criteria assess whether a model is equally accurate or well-calibrated across a set of predefined groups so that appropriate bias mitigation strategies can be implemented.
+{fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions and includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
 
@@
 
-<!---
-Jianhui - this is too similar to the original paper still and I suggest we emphasize that the package isn't restricted to biomedical data.
-We can of course use it as an example but this isn't the need for the package.
-Can you rewrite this to just be about the general use of fairness?
-i.e., point 1 - ml increasing integrated into different areas, point 2 - give a few examples, point 3 - need for the package in light of other work.
-I don't think interpetable is the correct word here, the key difference is allowing for uncertainty quantification, we also provide ratio + difference estimates, i dont think R fairness package does, I would emphasize
---->
+ML models are increasingly integrated into high-stakes domains to support decision making that significantly impacts individuals and society more broadly, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21].
+Mounting evidence suggest that these models often exhibit bias across groups defined by protected attributes.
+For example, within criminal justice, the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) software, a tool used by U.S.
+courts to evaluate the risk of defendants becoming recidivists, was found to incorrectly classify Black defendants as high-risk at nearly twice the rate of white defendants[@mattuMachineBias].
+This bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to white defendants with similar risk profiles.
+Similarly, within healthcare, a commercial risk-prediction algorithm deployed in the U.S.
+to identify patients with complex health needs for high-risk care management programs was shown to be significantly lower calibrated for Black patients relative to white patients[@obermeyerDissectingRacialBias2019].
+This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to white patients.
+These examples illustrate that there is an urgent need for ML practitioners and researchers to ensure that ML models support fair decision before they are deployed in real-world applications.
 
-Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21].
-As ML systems become a commonplace in decision-making processes, mounting evidence suggest that they often exhibit bias towards gender and ethnicity.
-For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants[@mattuMachineBias].
-As COMPAS was used in court to inform judicial decision-making, this bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to White defendants with similar risk profiles.
-Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients[@obermeyerDissectingRacialBias2019].
-This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to White patients.
-Therefore, there is a urgent need for companies, engineers, and researchers to ensure that ML models are fair and equitable across different groups before they are deployed in real-world applications.
-
-While existing software can compute fairness metrics, they provide only point estimates without quantifying the uncertainty around those metrics.
+While existing software can compute group fairness criteria, they only point estimates and/or visualizations without quantifying the uncertainty around the criteria.
 This limitation prevents users from determining whether observed disparities between groups are statistically significant or merely the result of random variation due to finite sample size, potentially leading to incorrect conclusions about fairness violations.
-The {fairmetrics} R package addresses this gap by providing bootstrap-based confidence intervals (CIs) for both difference-based and ratio-based group fairness metrics, enabling users to make statistically grounded decisions about the fairness of their models.
+The {fairmetrics} R package addresses this gap by providing bootstrap-based confidence intervals (CIs) for both difference-based and ratio-based group fairness metrics, empowering users to make statistically grounded decisions about the fairness of their models, which is inconsistently done in practice.
 

### 2025-05-30 paper.Rmd
Scope looks good.
Edit fairness criteria section.

@@
 
-<!---
-This section needs to be cleaned up for clarity - see comments below.
---->
-
-Group fairness criteria are typically classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022].
-Independence requires that the model’s predictions be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.
-Separation requires independence between the prediction and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within positive (or negative) outcome class.
+Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022].
+Independence requires that the model’s classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.
+Separation requires independence between the classifications and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within the positive (or negative) outcome class.
 Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model’s prediction is known, the protected attribute provides no additional information about the true outcome.
-
-<!---
-1.
-Jianhui - The scope of the package is not defined in the above paragprah so one would not understand the definitons below (i.e., "positive classification" is never defined).
-Specifically, you need to say you consider binary classification and a binary protected attribute - I do not believe your package handles more than.
-Explain why this is done and then update the definitions below to reflect this.
-
-2. Jianhui - why do you use bootstrap for this package rather than IF?  Is it because you don't have IF for all metrics?
-
-3. Jianhui - please update the initial sentences describing the 3 categories to be more intuitive, akin to what we have in the paper.  
---->
+Below we summarize the fairness metrics that are available within the {fairmetrics} package.
 
@@
 
--   **Statistical Parity:** Compares the overall rate of positive predictions between groups, irrespective of the true outcome.
+-   **Statistical Parity:** Compares the overall rate of positive predictions between groups.
 
@@
 
-<!---
-Use consistent language throughout - alternating between checks/compares/focuses/assess/measure/etc is confusing as all functions do the same thing.  Suggest use "Compares" for all.
-
-Ben Response: Rewritten to use "Compares".
---->
-
 ## Separation
 
--   **Equal Opportunity:**  Compares disparities in false negative rates between groups, quantifying any difference in missed positive cases.
+-   **Equal Opportunity:**  Compares disparities in the false negative rates between groups, quantifying differences in the likelihood of missing positive outcomes.
 
-<!---
-Do you need the acronyms for FNR and FPR? If you don't use them later, they don't need to be defined.
+-   **Predictive Equality:** Compares the false positive rates (FPR) between groups, quantifying differences in the likelihood of incorrectly labeling negative outcomes as positive.
 
-Ben Response: FPR is used. FNR is not. Removed FNR. 
---->
--   **Predictive Equality:** Compares false positive rates (FPR) between groups, ensuring that no group is disproportionately flagged as positive when the true outcome is negative.
+-   **Balance for Positive Class:**  Compares the average of the predicted probabilities among individuals whose true outcome is positive across groups.
 
-<!---
-Updated the two defs below. It only checks the first-moment of the predicted probabilities, does not check the distribution.
-
-Also updated the name, Balance of Positive/Negative Class is the name the original paper used.
---->
-
--   **Balance for Positive Class:**  Compares if the average of predicted probabilities among individuals whose true outcome is positive is similar between groups.
-
--   **Balance for Negative Class:** Compares if the average of predicted probabilities among individuals whose true outcome is negative is similar between groups.
+-   **Balance for Negative Class:** Compares the average of the predicted probabilities among individuals whose true outcome is negative across groups.
 
@@
 
--   **Predictive Parity:** Compares positive predictive values across groups, assessing whether the precision of positive predictions is equivalent.
+-   **(Positive) Predictive Parity:** Compares the positive predictive values across groups, assessing differences in the precision of positive predictions.
+
+-   **(Negative) Predictive Parity:** Compares the negative predictive values across groups, assessing differences in the precision of negative predictions. [Note: Please add this to the package.]
 
@@
 
--   **Brier Score Parity:** Compares the Brier score—the mean squared error of predicted probabilities—is similar across groups, indicating comparable calibration.
+-   **Brier Score Parity:** Compares the Brier score (i.e., the mean squared error of predicted probabilities) across groups, evaluating differences in calibration.
 
--   **Accuracy Parity:** Compares the overall accuracy of a predictive model is equivalent across groups.
+-   **Accuracy Parity:** Compares the overall accuracy of a predictive model across groups.
 

### 2025-05-30 paper.Rmd
Update evaluating fairness criteria section.

@@
 
--   **Treatment Equality:** Compares the ratio of false negatives to false positives across groups, ensuring the balance of missed detections versus false alarms is consistent.
+-   **Treatment Equality:** Compares the ratio of false negatives to false positives across groups, evaluating whether the trade-off between missed detections of positive outcomes and false alarms of negative outcomes is balanced.
 
@@
 
-<!--
-1.
-Ben - The package doesn't require data splitting, some people may want to use the package and incorporate CV within training.
-Also, sometimes people evaluate fairness metrics on validation data.
-So suggest you simplify text + figure.
-Just start with the input.
-Also, it doesn't seem like multiple should be a separate node?
-doesnt it just include all the 4 boxes (sep, suff, ind, other)?
-can this be depicted?
-Also suggest you make the names more descriptive "Separation-based Metrics", "Independence-based Metrics", etc.
-
-Ben Response: Created 3 different figures to review. I think v3 is the best. Please critique if there are any issues.
---->
-The primary input to the {fairmetrics} package is a data frame or tibble which containing the model's predictions, true outcomes, and the protected attribute in question.
+<!-- The figure looks good but do not use any abbreviations (i.e., Predicted probabilities).
+Also, we use "protected" not sensitive attribute in paper so update.
+Also add negative predictive parity once added to the package.
+For combined: What's the difference between get_all and get_fairness?
+-->
+The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest.
 \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}.
-It is possible to evaluate a model for a specific or multiple group fairness metrics.
+Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function..
 
@@
 
-<!---
-Previously edited by Jesse, now edited by Ben
---->
-A simple example of how to use the {fairmetrics} package is shown below.
-The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data].
-This dataset consists of 1776 hemodynamically stable patients with respiratory failure, and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating whether the patient died within 28 days of admission.
+A simple example of how to use the {fairmetrics} package is illustrated below.
+The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from the MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in the {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data].
+The dataset consists of 1,776 hemodynamically stable patients with respiratory failure and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating whether the patient died within 28 days of admission.
 
-While the choice of fairness metric used is context dependent, we show all metrics available with the `get_fairness_metrics()` function.
+While the choice of fairness metric used is context dependent, we show all metrics available with the `get_fairness_metrics()` function for the purposes of illustration.
 In this example, we evaluate the model's fairness with respect to the protected attribute `gender`.
 For conditional statistical parity, we condition on patients older than 60 years old.
-The model is trained on a subset of the data, and predictions are made on a test set.
-<!---
-1. Ben - need to add detail on what the detail is in the above paragraph. What does the data contaon, where did it come from, what is the sample size, etc.
-
-Ben Response: Done.
-
-2.
-Ben - suggest you show more of the functions in the package - say that, while the choice of metric is context dependent, we show all metrics to show the full range of the package.
-Also include interpretation of the various results.
-Simple way to do this is show, the multiple output then show an example of one of the separation-based criterion or other to show users they can specify what they want.
-
-Ben Response: Please see below for the examples and review.
----->
+The model is trained on a subset of the data and the predictions are made and evaluated on a test set.
 

### 2025-05-30 paper.Rmd
Add more comments to the real data example.

@@
 <!-- The figure looks good but do not use any abbreviations (i.e., Predicted probabilities).
-Also, we use "protected" not sensitive attribute in paper so update.
+Also, we use "protected" not sensitive attribute in paper so needs an update.
 Also add negative predictive parity once added to the package.
 For combined: What's the difference between get_all and get_fairness?
+This isn't clear to me.
+Also I think you should have an argument for level for the CIs - that is, users input 0.95 for 95\% or 0.99 etc as that should be adjustable.
 -->
@@
 
-While the choice of fairness metric used is context dependent, we show all metrics available with the `get_fairness_metrics()` function for the purposes of illustration.
+While the choice of fairness criteria used is context dependent, we show all criteria available with the `get_fairness_metrics()` function for the purposes of illustration.
 In this example, we evaluate the model's fairness with respect to the protected attribute `gender`.
@@
 The model is trained on a subset of the data and the predictions are made and evaluated on a test set.
+The `get_fairness_metrics()` function outputs difference and ratio-based metrics as well as their corresponding confidence intervals.
+A statistically significant difference across groups at a given level of significance is indicated when the confidence interval for a difference-based metric does not include zero or when the interval for a ratio-based metric does not include one.
 
@@
 
-Among the fairness metrics calculated, metrics whose difference confidence intervals cross zero and whose ratio confidence intervals cross one indicate no significant difference between the groups.
-In this example, the statistical parity, conditional statistical pairty, equal opportunity, predictive equality, positive class balance, negative class balance, and treatment equaltiy metrics show a significant differences.
-While, the predictive parity, Brier score parity, and overall accuracy parity metrics do not show significant differences between the groups.
-
 Should the user wish to calculate an individual metric, it is possible to use any of the `eval_*` functions.
-For example, to calculate the equal opportunity metric, the user can use the `eval_equal_opportunity()` function.
+For example, to calculate the equal opportunity metric, the user can call the `eval_equal_opportunity()` function.
 
@@
 This function returns a data frame with the metrics calculated for each group.
+[ NOTE: To me it is unclear why this is a separate function.
+I think instead it makes sense to have an argument in your get_fairness_metrics function that turns the CI calculation on or off.
+]
 
@@
 
-<!---
-## Both - you are missing this one: https://mlr3fairness.mlr-org.com
-## Do these metrics give difference and ratio based criteria? Are there any other differences?
-
-Ben Response: Added mlr3fairness package updated the differences.
-
-For comparison:
-
-- {fairness}- does not give difference or ratio based criteria, only point estimates. The package also produces plots which {fairmetrics} does not (not in our scope).
-- {fairmodels}- does give ratio criteria but not GROUP ratio based criteria, only overall ratio.
-The package also produces plots which {fairmetrics} does not (not in our scope).
-No difference based criteria.
-- Same with {mlr3fairness} - does not give difference or ratio based criteria, only point estimates. The package also produces plots which {fairmetrics} does not (not in our scope).
-
-(I would also add that the {fairmodels} has added complexity required to using it, and {mlr3fairness} packages are designed for use with the mlr3 framework, which is not as widely adopted in the R community.
-This could be a "skill issue" on my part.
-However I don't think such a comment would be appropriate for the paper.)
-
---->
 Other R packages similar to {fairmetrics} include {fairness}[@fairness_package], {fairmodels} [@wisniewski2022fairmodels] and {mlr3fairness}[mlr3fairness_package].

### 2025-05-30 paper.Rmd
Finish adding comments.

@@
 The differences between {fairmetrics} and these other packages is twofold.
-The primary difference between is that {fairmetrics} calculates the ratio and difference between group fairness criterion and allows estimated confidence intervals of fairness metrics via bootstrap - allowing for more meaningful inferences about the fairness metrics calculated.
+The primary difference between is that {fairmetrics} calculates the ratio and difference between group fairness criterion and their corresponding confidence intervals of fairness metrics via bootstrap, allowing for more meaningful inferences about the fairness criteria.
 Additionally, in contrast to the {fairmodels}, {fairness} and {mlr3fairness} packages, the {fairmetrics} package does not posses any external dependencies and has a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems.
@@
 \end{tabular}
-\caption{Memory usage and dependencies of {fairmetrics} vs similar packages (MB)}
+\caption{Memory usage (in MB) and dependencies of {fairmetrics} vs similar packages.}
 \label{tab:memory_dep_usage}
@@
 
-For python users, the {fairlearn} library [@fairlearn_paper] provides a broader set of fairness metrics and algorithms.
+For python users, the {fairlearn} library [@fairlearn_paper] provides additional fairness metrics and algorithms.
 The {fairmetrics} package is designed for seemless integration with R workflows, making it a more convenient choice for R-based ML applications.

### 2025-05-30 paper.Rmd
Add one comment.

@@
 
-Should the user wish to calculate an individual metric, it is possible to use any of the `eval_*` functions.
-For example, to calculate the equal opportunity metric, the user can call the `eval_equal_opportunity()` function.
+Should the user wish to calculate an individual criteria, it is possible to use any of the `eval_*` functions.
+For example, to calculate equal opportunity, the user can call the `eval_equal_opportunity()` function.
 
@@
 This function returns a data frame with the metrics calculated for each group.
-[ NOTE: To me it is unclear why this is a separate function.
-I think instead it makes sense to have an argument in your get_fairness_metrics function that turns the CI calculation on or off.
-]
+
+[ NOTE: To me it is unclear why this is a separate function.  I think instead it makes sense to have an argument in your get_fairness_metrics function that turns the CI calculation on or off.  ]
 

### 2025-06-01 R/Fairness.R
adding negative pred parity, renaming eval_pred_parity to eval_pos_pred_parity

@@ -584,9 +584,9 @@ eval_cond_stats_parity <- function(data, outcome, group,
   }
 }
 
-#' Examine Predictive Parity of a Model
+#' Examine Positive Predictive Parity of a Model
 #'
-#' This function evaluates *predictive parity (PP)*, a key fairness criterion that
+#' This function evaluates *positive predictive predictive parity*, a key fairness criterion that
 #' compares the *Positive Predictive Value (PPV)* between groups defined by a sensitive attribute.
 #' In other words, it assesses whether, among individuals predicted to be positive,
 #' the probability of being truly positive is equal across subgroups.
@@ -638,8 +638,8 @@ eval_cond_stats_parity <- function(data, outcome, group,
 #' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
-#' # Evaluate Predictive Parity
-#' eval_pred_parity(
+#' # Evaluate Positive Predictive Parity
+#' eval_pos_pred_parity(
 #'   data = test_data,
 #'   outcome = "day_28_flg",
 #'   group = "gender",
@@ -647,9 +647,10 @@ eval_cond_stats_parity <- function(data, outcome, group,
 #'   cutoff = 0.41
 #' )
 #' }
+#' @seealso \code{\link{eval_neg_pred_parity}}
 #' @export
 
-eval_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
+eval_pos_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                              bootstraps = 2500, alpha = 0.05,
                              digits = 2, message = TRUE) {
   # Check if outcome is binary
@@ -706,15 +707,149 @@ eval_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint 
 
   if (message) {
     if (lower_ci > 0 || upper_ci < 0) {
-      cat("There is evidence that model does not satisfy predictive parity.\n")
+      cat("There is evidence that model does not satisfy positive predictive parity.\n")
     } else {
-      cat("There is not enough evidence that the model does not satisfy
-            predictive parity.\n")
+      cat("There is not enough evidence that the model does not satisfy positive predictive parity.\n")
[…]

### 2025-06-01 R/Fairness.R
getting CRAN ready

@@ -912,7 +912,7 @@ eval_neg_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, conf
 #'   cutoff = 0.41
 #' )
 #' }
-#' @seealso \code{\link{eval_pred_parity}}, \code{\link{eval_stats_parity}}
+#' @seealso \code{\link{eval_pos_pred_parity}},  \code{\link{eval_neg_pred_parity}}, \code{\link{eval_stats_parity}}
 #' @export
 
 eval_pred_equality <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
@@ -1326,7 +1326,7 @@ eval_acc_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint =
 #' )
 #' }
 #'
-#' @seealso \code{\link{eval_acc_parity}}, \code{\link{eval_cond_acc_equality}}, \code{\link{eval_pred_parity}}
+#' @seealso \code{\link{eval_acc_parity}}, \code{\link{eval_cond_acc_equality}}, \code{\link{eval_pos_pred_parity}},  \code{\link{eval_neg_pred_parity}}
 #' @export
 
 eval_bs_parity <- function(data, outcome, group, probs, confint = TRUE,
@@ -1465,7 +1465,7 @@ eval_bs_parity <- function(data, outcome, group, probs, confint = TRUE,
 #'   message = FALSE
 #' )
 #' }
-#' @seealso \code{\link{eval_acc_parity}}, \code{\link{eval_bs_parity}}, \code{\link{eval_pred_parity}}
+#' @seealso \code{\link{eval_acc_parity}}, \code{\link{eval_bs_parity}}, \code{\link{eval_pos_pred_parity}},  \code{\link{eval_neg_pred_parity}}
 #' @export
 
 eval_treatment_equality <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,

### 2025-06-01 R/get_fairness_metrics.R

@@ -172,7 +172,19 @@ get_fairness_metrics <- function(data,
     message = FALSE
   )
 
-  pred_parity <- eval_pred_parity(
+  pos_pred_parity <- eval_pos_pred_parity(
+    data = data,
+    outcome = outcome,
+    group = group,
+    probs = probs,
+    cutoff = cutoff,
+    bootstraps = bootstraps,
+    alpha = alpha,
+    digits = digits,
+    message = FALSE
+  )
+
+  neg_pred_parity <- eval_neg_pred_parity(
     data = data,
     outcome = outcome,
     group = group,
@@ -223,7 +235,8 @@ get_fairness_metrics <- function(data,
   pred_equality[["Metric"]] <- "Predictive Equality"
   pos_class_bal[["Metric"]] <- "Balance for Positive Class"
   neg_class_bal[["Metric"]] <- "Balance for Negative Class"
-  pred_parity[["Metric"]] <- "Predictive Parity"
+  pos_pred_parity[["Metric"]] <- "Positive Predictive Parity"
+  neg_pred_parity[["Metric"]] <- "Negative Predictive Parity"
   bs_parity[["Metric"]] <- "Brier Score Parity"
   acc_parity[["Metric"]] <- "Overall Accuracy Parity"
   treatment_equality[["Metric"]] <- "Treatment Equality"
@@ -238,7 +251,8 @@ get_fairness_metrics <- function(data,
         pred_equality,
         pos_class_bal,
         neg_class_bal,
-        pred_parity,
+        pos_pred_parity,
+        neg_pred_parity,
         bs_parity,
         acc_parity,
         treatment_equality
@@ -254,7 +268,8 @@ get_fairness_metrics <- function(data,
         pred_equality,
         pos_class_bal,
         neg_class_bal,
-        pred_parity,
+        pos_pred_parity,
+        neg_pred_parity,
         bs_parity,
         acc_parity,
         treatment_equality

### 2025-06-01 R/Fairness.R
removing get_all_metrics and having get_fairness_metrics cover it

@@ -17,6 +17,7 @@
 #' predicted probabilities or risk scores.
 #' @param cutoff A numeric value used to threshold predicted probabilities into
 #' binary decisions; defaults to 0.5.
+#' @param confint Whether to compute 95% confidence interval, default is TRUE.
 #' @param bootstraps An integer specifying the number of bootstrap resamples for
 #' constructing confidence intervals; defaults to 2500.
 #' @param alpha Significance level for constructing the (1 - \code{alpha})
@@ -77,7 +78,7 @@
 #' }
 #' @export
 
-eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5,
+eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5,confint = TRUE,
                         bootstraps = 2500, alpha = 0.05, digits = 2,
                         message = TRUE) {
   # Check if outcome is binary
@@ -94,48 +95,63 @@ eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5,
   fnr_diff <- fnr[[1]] - fnr[[2]]
   fnr_ratio <- fnr[1] / fnr[[2]]
 
-  # Calculate difference confidence interval
-  se <- replicate(bootstraps, {
-    # Bootstrap within each group
-    group1 <- sample(which(data[[group]] == unique(data[[group]])[1]),
-      replace = TRUE
+  if(confint){
+    # Calculate difference confidence interval
+    se <- replicate(bootstraps, {
+      # Bootstrap within each group
+      group1 <- sample(which(data[[group]] == unique(data[[group]])[1]),
+                       replace = TRUE
+      )
+      group2 <- sample(which(data[[group]] == unique(data[[group]])[2]),
+                       replace = TRUE
+      )
+      data_boot <- rbind(data[group1, ], data[group2, ])
+
+      fnr_boot <- 1 - get_tpr(
+        data = data_boot, outcome = outcome, group = group,
+        probs = probs, cutoff = cutoff
+      )
+      return(c(fnr_boot[[1]] - fnr_boot[[2]], log(fnr_boot[[1]] / fnr_boot[[2]])))
+    })
+
[…]

### 2025-06-01 R/ModelPerformance.R

@@ -218,83 +218,3 @@ get_avg_prob <- function(data, outcome, group, probs, digits = 2) {
   }
   return(avg_prob)
 }
-
-#' Calculate the all metrics at once
-#'
-#' This function computes a comprehensive set of fairness-related performance metrics across the
-#' levels of a sensitive attribute. It includes standard classification metrics (e.g., TPR, FPR, PPV, NPV)
-#' as well as fairness-specific indicators like predicted positive rates and error ratios.
-#'
-#' This is useful for quickly assessing multiple fairness dimensions of a binary classifier in one step.
-#'
-#' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
-#' @param outcome the name of the outcome variable, it must be binary
-#' @param group the name of the sensitive attribute
-#' @param probs the name of the predicted outcome variable
-#' @param cutoff the threshold for the predicted outcome, default is 0.5
-#' @param digits the number of digits to round the result to, default is 2
-#' @return a Data frame of all metrics
-#' @examples
-#' \donttest{
-#' library(fairmetrics)
-#' library(dplyr)
-#' library(magrittr)
-#' library(randomForest)
-#' data("mimic_preprocessed")
-#' set.seed(123)
-#' train_data <- mimic_preprocessed %>%
-#'   dplyr::filter(dplyr::row_number() <= 700)
-#' # Fit a random forest model
-#' rf_model <- randomForest::randomForest(factor(day_28_flg) ~ ., data = train_data, ntree = 1000)
-#' # Test the model on the remaining data
-#' test_data <- mimic_preprocessed %>%
-#'   dplyr::mutate(gender = ifelse(gender_num == 1, "Male", "Female"))%>%
-#'   dplyr::filter(dplyr::row_number() > 700)
-#'
-#' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
-#'
-#' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
-#' # We choose threshold = 0.41 so that the overall FPR is around 5%.
-#'
-#' # Calculate All Metrics
-#' get_all_metrics(
-#'   dat = test_data,
-#'   outcome = "day_28_flg",
-#'   group = "gender",
-#'   probs = "pred",
[…]

### 2025-06-01 R/get_fairness_metrics.R

@@ -13,7 +13,8 @@
 #' - **Balance for Positive Class**: Checks whether the predicted probability distributions for
 #'   positive outcomes are similar across groups.
 #' - **Balance for Negative Class**: Same as above, but for negative outcomes.
-#' - **Predictive Parity**: Difference in positive predictive values (precision) across groups.
+#' - **Positive Predictive Parity**: Difference in positive predictive values (precision) across groups.
+#' - **Negative Predictive Parity**: Difference in negative predictive values across groups.
 #' - **Brier Score Parity**: Difference in Brier scores across groups.
 #' - **Overall Accuracy Parity**: Difference in overall accuracy across groups.
 #' - **Treatment Equality**: Ratio of false negatives to false positives across groups.
@@ -28,6 +29,7 @@
 #' the sign of the condition and the value to threshold the continuous variable
 #' (e.g. "<50", ">50", "<=50", ">=50").
 #' @param probs The name of the column with predicted probabilities.
+#' @param confint Logical indicating whether to calculate confidence intervals.
 #' @param cutoff Numeric threshold for classification. Default is 0.5.
 #' @param bootstraps Number of bootstrap samples. Default is 2500.
 #' @param alpha Significance level for confidence intervals. Default is 0.05.
@@ -66,7 +68,9 @@
 #'  group2 = "age",
 #'  condition = ">=60",
 #'  probs = "pred",
-#'  cutoff = 0.41
+#'  confint = TRUE,
+#'  cutoff = 0.41,
+#'  alpha = 0.05
 #' )
 #' }
 #'
@@ -78,6 +82,7 @@ get_fairness_metrics <- function(data,
                                  group2 = NULL,
                                  condition = NULL,
                                  probs,
+                                 confint = TRUE,
                                  cutoff = 0.5,
                                  bootstraps = 2500,
                                  alpha = 0.05,
@@ -87,6 +92,7 @@ get_fairness_metrics <- function(data,
     outcome = outcome,
     group = group,
     probs = probs,
+    confint = confint,
     cutoff = cutoff,
     bootstraps = bootstraps,
     alpha = alpha,
@@ -106,6 +112,7 @@ get_fairness_metrics <- function(data,
       group2 = group2,
       condition = condition,
       probs = probs,
[…]

### 2025-06-01 tests/testthat.R

@@ -8,4 +8,4 @@
 
 library(testthat)
 
-test_check("FairnessEval")
+test_check("fairmetrics")

### 2025-06-01 tests/testthat/helper.R

@@ -1,4 +1,4 @@
-library(FairnessEval)
+library(fairmetrics)
 library(dplyr)
 library(randomForest)
 # Data for tests

### 2025-06-01 tests/testthat/test-ModelPerformance.R

@@ -1,15 +1,0 @@
-test_that('get_all_metrics test', {
-  expect_no_error({
-    source("helper.R")
-    get_all_metrics(
-      dat = test_data,
-      outcome = "day_28_flg",
-      group = "gender",
-      probs = "pred",
-      cutoff = 0.41,
-      digits = 2
-    )
-  })
-
-})
-

### 2025-06-01 tests/testthat/test-get_fairness_metrics.R

@@ -1,17 +1,38 @@
-test_that('get_fairness_metrics test', {
+test_that('get_fairness_metrics CI test', {
   expect_no_error({
     source("helper.R")
     get_fairness_metrics(
-      dat = test_data,
+      data = test_data,
       outcome = "day_28_flg",
       group = "gender",
       group2 = "age",
-      condition = ">= 60",
+      condition = ">=60",
       probs = "pred",
+      confint = TRUE,
       cutoff = 0.41,
       bootstraps = 2500,
       alpha = 0.05,
-      digits =2
+      digits = 2
+    )
+
+  })
+})
+
+test_that('get_fairness_metrics no CI test', {
+  expect_no_error({
+    source("helper.R")
+    get_fairness_metrics(
+      data = test_data,
+      outcome = "day_28_flg",
+      group = "gender",
+      group2 = "age",
+      condition = ">=60",
+      probs = "pred",
+      confint = FALSE,
+      cutoff = 0.41,
+      bootstraps = 2500,
+      alpha = 0.05,
+      digits = 2
     )
 
   })

### 2025-06-01 paper.Rmd
Update paper.Rmd
current paper

@@
   rticles::joss_article
+  #rticles::arxiv_article
   # md_document:
@@
 Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022].
-Independence requires that the model’s classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.
+Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.
 Separation requires independence between the classifications and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within the positive (or negative) outcome class.
-Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model’s prediction is known, the protected attribute provides no additional information about the true outcome.
+Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model's prediction is known, the protected attribute provides no additional information about the true outcome.
 Below we summarize the fairness metrics that are available within the {fairmetrics} package.
@@
 
--   **(Positive) Predictive Parity:** Compares the positive predictive values across groups, assessing differences in the precision of positive predictions.
+-   **Positive Predictive Parity:** Compares the positive predictive values across groups, assessing differences in the precision of positive predictions.
 
--   **(Negative) Predictive Parity:** Compares the negative predictive values across groups, assessing differences in the precision of negative predictions. [Note: Please add this to the package.]
+-   **Negative Predictive Parity:** Compares the negative predictive values across groups, assessing differences in the precision of negative predictions.
 
@@
 Also I think you should have an argument for level for the CIs - that is, users input 0.95 for 95\% or 0.99 etc as that should be adjustable.
+
+BEN RESPONSE:
+
+1. Updated the visual to not use abbreviations.
+2. Changed "sensitive' attribute to "protected" attribute
+3. Added negative predictive parity to package (`eval_neg_pred_parity()`; renamed `eval_pred_parity()` to `eval_pos_pred_parity()`). Submitted & Uploaded to CRAN. Added to the visual as well.
+4.  Noted. Refactored the code to allow get_fairness_metrics() to cover the functionality of get_all_metrics(). This can be specified with the confint = FALSE.
+5.  There is an argument level for the CIs, its the `alpha` parameter which by default is set to 0.05. Added the argument to the code. Do you want me to talk about it in the text as well? 
  -->
@@
 \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}.
-Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function..
+Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function.
 
-![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow-v3.png)
+![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow.png)
 
@@
 
-# Evaluate predictive equality
-# (Setting message=FALSE to avoid cluttering the output)
-
+# Get fairness metrics
+# Setting alpha=0.05 for 95% confidence intervals
 get_fairness_metrics(
@@
  probs = "pred",
- cutoff = 0.41
+ cutoff = 0.41, 
+ alpha = 0.05
 )
 
-#>                                       Metric GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#> 1                         Statistical Parity        0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
-#> 2  Conditional Statistical Parity (age >=60)        0.34      0.21       0.13   [0.05, 0.21]  1.62 [1.18, 2.22]
-#> 3                          Equal Opportunity        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
-#> 4                        Predictive Equality        0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.39, 5.1]
-#> 5                 Balance for Positive Class        0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.41]
-#> 6                 Balance for Negative Class        0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
-#> 7                          Predictive Parity        0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.23]
-#> 8                         Brier Score Parity        0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
-#> 9                    Overall Accuracy Parity        0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
-#> 10                        Treatment Equality        1.03      3.24      -2.21 [-4.35, -0.07]  0.32 [0.15, 0.68]
+#>                 Metric                          Full Metric Name GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
+#> 1                   PPR                        Statistical Parity        0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
+#> 2                   PPR Conditional Statistical Parity (age >=60)        0.34      0.21       0.13   [0.05, 0.21]  1.62 [1.18, 2.22]
+#> 3                   FNR                         Equal Opportunity        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
+#> 4                   FPR                       Predictive Equality        0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.39, 5.1]
+#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.41]
+#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
+#> 7                   PPV                Positive Predictive Parity        0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.23]
+#> 8                   NPV                Negative Predictive Parity        0.92      0.90       0.02  [-0.15, 0.19]  1.02 [0.79, 1.33]
+#> 9           Brier Score                        Brier Score Parity        0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.88, 1.43]
+#> 10             Accuracy                   Overall Accuracy Parity        0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
+#> 11          FN/FP Ratio                        Treatment Equality        1.03      3.24      -2.21   [-4.5, 0.08]  0.32 [0.15, 0.69]
 ```
@@
   probs = "pred",
-  cutoff = 0.41
+  confint = TRUE,
+  cutoff = 0.41,
+  alpha = 0.05
 )
@@
 #>1    FNR        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.85]
-```
-
-For group-specific performance metrics, such as the True Positive Rate (TPR), False Positive Rate (FPR), Positive Predictive Value (PPV), Negative Predictive Value (NPV), and others, the `get_all_metrics()` function can be used.
-This function returns a data frame with the metrics calculated for each group.
-
-[ NOTE: To me it is unclear why this is a separate function.  I think instead it makes sense to have an argument in your get_fairness_metrics function that turns the CI calculation on or off.  ]
-
-```r
-get_all_metrics(
-  dat = test_data,
-  outcome = "day_28_flg",
-  group = "gender",
-  probs = "pred",
-  cutoff = 0.41
-)
-
-#>         Metric Group Female Group Male
-#> 1           TPR         0.62       0.38
-#> 2           FPR         0.08       0.03
-#> 3           PPR         0.17       0.08
-#> 4           PPV         0.62       0.66
-#> 5           NPV         0.92       0.90
-#> 6           ACC         0.87       0.88
-#> 7   Brier Score         0.09       0.08
-#> 8   FN/FP Ratio         1.03       3.24
-#> 9 Avg Pred Prob         0.21       0.14
 ```

### 2025-06-02 paper.Rmd
Edit summary.

@@
 
-Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for particular groups, particularly those defined by protected attributes such as race, gender, or age.
+Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for specific groups, particularly those defined by protected attributes such as race, gender, or age.
 Evaluating fairness is a critical aspect of ML model development, as biased models can perpetuate structural inequalities.
@@
 # Evaluating Fairness Criteria
-
-<!-- The figure looks good but do not use any abbreviations (i.e., Predicted probabilities).
-Also, we use "protected" not sensitive attribute in paper so needs an update.
-Also add negative predictive parity once added to the package.
-For combined: What's the difference between get_all and get_fairness?
-This isn't clear to me.
-Also I think you should have an argument for level for the CIs - that is, users input 0.95 for 95\% or 0.99 etc as that should be adjustable.
-
-BEN RESPONSE:
-
-1. Updated the visual to not use abbreviations.
-2. Changed "sensitive' attribute to "protected" attribute
-3. Added negative predictive parity to package (`eval_neg_pred_parity()`; renamed `eval_pred_parity()` to `eval_pos_pred_parity()`). Submitted & Uploaded to CRAN. Added to the visual as well.
-4.  Noted. Refactored the code to allow get_fairness_metrics() to cover the functionality of get_all_metrics(). This can be specified with the confint = FALSE.
-5.  There is an argument level for the CIs, its the `alpha` parameter which by default is set to 0.05. Added the argument to the code. Do you want me to talk about it in the text as well? 
- -->
 The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest.

### 2025-06-02 paper.Rmd
Add note with additional citations.

@@
     index: 1
-date: "2025-03-08"
+date: "2025-06-02"
 bibliography: paper.bib
@@
 Group-based fairness criteria assess whether a model is equally accurate or well-calibrated across a set of predefined groups so that appropriate bias mitigation strategies can be implemented.
-{fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions and includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
+{fairmetrics} provides both point and interval estimates for multiple metrics through a convenient wrapper function and includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
 
@@
 Similarly, within healthcare, a commercial risk-prediction algorithm deployed in the U.S.
-to identify patients with complex health needs for high-risk care management programs was shown to be significantly lower calibrated for Black patients relative to white patients[@obermeyerDissectingRacialBias2019].
+to identify patients with complex health needs for high-risk care management programs was shown to be significantly less calibrated for Black patients relative to white patients[@obermeyerDissectingRacialBias2019].
 This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to white patients.
-These examples illustrate that there is an urgent need for ML practitioners and researchers to ensure that ML models support fair decision before they are deployed in real-world applications.
+These examples illustrate that there is an urgent need for practitioners and researchers to ensure that ML models support fair decision making before they are deployed in real-world applications.
 
@@
 Second, group fairness is the most widely used framework for binary classification tasks[@mehrabi_survey_21].
-Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness.
+Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness[NOTE:citation to add: https://arxiv.org/abs/2205.05770].
 This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.
@@
 
-Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022].
+Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022][NOTE: cite https://arxiv.org/abs/2406.09307].
 Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.

### 2025-06-02 paper.Rmd
Fix typo.

@@
 
--   **Brier Score Parity:** Compares the Brier score (i.e., the mean squared error of predicted probabilities) across groups, evaluating differences in calibration.
+-   **Brier Score Parity:** Compares the Brier score (i.e., the mean squared error of the predicted probabilities) across groups, evaluating differences in calibration.
 

### 2025-06-02 paper.Rmd
Update paper.Rmd

@@
 The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from the MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in the {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data].
-The dataset consists of 1,776 hemodynamically stable patients with respiratory failure and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating whether the patient died within 28 days of admission.
+The dataset consists of 1,776 hemodynamically stable patients with respiratory failure and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating whether the patient died within 28 days of hospital admission.
 

### 2025-06-02 paper.Rmd
add space before citation

@@
 
-ML models are increasingly integrated into high-stakes domains to support decision making that significantly impacts individuals and society more broadly, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21].
+ML models are increasingly integrated into high-stakes domains to support decision making that significantly impacts individuals and society more broadly, including criminal justice, healthcare, finance, employment, and education [@mehrabi_survey_21].
 Mounting evidence suggest that these models often exhibit bias across groups defined by protected attributes.
 For example, within criminal justice, the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) software, a tool used by U.S.
-courts to evaluate the risk of defendants becoming recidivists, was found to incorrectly classify Black defendants as high-risk at nearly twice the rate of white defendants[@mattuMachineBias].
+courts to evaluate the risk of defendants becoming recidivists, was found to incorrectly classify Black defendants as high-risk at nearly twice the rate of white defendants [@mattuMachineBias].
 This bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to white defendants with similar risk profiles.
 Similarly, within healthcare, a commercial risk-prediction algorithm deployed in the U.S.
-to identify patients with complex health needs for high-risk care management programs was shown to be significantly less calibrated for Black patients relative to white patients[@obermeyerDissectingRacialBias2019].
+to identify patients with complex health needs for high-risk care management programs was shown to be significantly less calibrated for Black patients relative to white patients [@obermeyerDissectingRacialBias2019].
 This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to white patients.
@@
 This restriction reflects standard practice in the fairness literature and is motivated by several considerations.
-First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative[@mehrabi_survey_21].
-Second, group fairness is the most widely used framework for binary classification tasks[@mehrabi_survey_21].
-Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness[NOTE:citation to add: https://arxiv.org/abs/2205.05770].
+First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative [@mehrabi_survey_21].
+Second, group fairness is the most widely used framework for binary classification tasks [@mehrabi_survey_21].
+Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness [@lum_bias_22].
 This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.
@@
 
-Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022][NOTE: cite https://arxiv.org/abs/2406.09307].
+Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022, @Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
 Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.
@@
 
-Other R packages similar to {fairmetrics} include {fairness}[@fairness_package], {fairmodels} [@wisniewski2022fairmodels] and {mlr3fairness}[mlr3fairness_package].
+Other R packages similar to {fairmetrics} include {fairness} [@fairness_package], {fairmodels} [@wisniewski2022fairmodels] and {mlr3fairness} [@mlr3fairness_package].
 The differences between {fairmetrics} and these other packages is twofold.

### 2025-06-04 paper.Rmd
fixing table format (reran full example)

@@
 #>                  Metric                          Full Metric Name GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#> 1                   PPR                        Statistical Parity        0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
-#> 2                   PPR Conditional Statistical Parity (age >=60)        0.34      0.21       0.13   [0.05, 0.21]  1.62 [1.18, 2.22]
-#> 3                   FNR                         Equal Opportunity        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
-#> 4                   FPR                       Predictive Equality        0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.39, 5.1]
-#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.41]
-#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
-#> 7                   PPV                Positive Predictive Parity        0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.23]
-#> 8                   NPV                Negative Predictive Parity        0.92      0.90       0.02  [-0.15, 0.19]  1.02 [0.79, 1.33]
-#> 9           Brier Score                        Brier Score Parity        0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.88, 1.43]
-#> 10             Accuracy                   Overall Accuracy Parity        0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
-#> 11          FN/FP Ratio                        Treatment Equality        1.03      3.24      -2.21   [-4.5, 0.08]  0.32 [0.15, 0.69]
+#> 1                   PPR                        Statistical Parity        1.12      1.06       0.06    [0.02, 0.1]  1.06 [1.02, 1.09]
+#> 2                   PPR Conditional Statistical Parity (age >=60)        1.23      1.16       0.07      [0, 0.14]  1.06    [1, 1.13]
+#> 3                   FNR                         Equal Opportunity       -0.35     -0.23      -0.12  [-0.26, 0.02]  1.52  [0.9, 2.58]
+#> 4                   FPR                       Predictive Equality        1.07      1.03       0.04   [0.01, 0.07]  1.04 [1.01, 1.07]
+#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.50      0.50       0.00         [0, 0]  1.00       [1, 1]
+#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.50      0.50       0.00         [0, 0]  1.00       [1, 1]
+#> 7                   PPV                Positive Predictive Parity        0.21      0.16       0.05       [0, 0.1]  1.31 [0.99, 1.74]
+#> 8                   NPV                Negative Predictive Parity        0.87      0.88      -0.01  [-0.06, 0.04]  0.99 [0.74, 1.32]
+#> 9           Brier Score                        Brier Score Parity        0.37      0.40      -0.03 [-0.04, -0.02]  0.92 [0.89, 0.96]
+#> 10             Accuracy                   Overall Accuracy Parity        1.00      1.01      -0.01  [-0.05, 0.03]  0.99 [0.95, 1.03]
+#> 11          FN/FP Ratio                        Treatment Equality        0.13      0.12       0.01  [-0.04, 0.06]  1.08  [0.73, 1.6]
 ```

### 2025-06-04 paper.Rmd
current edits

@@
 
-While existing software can compute group fairness criteria, they only point estimates and/or visualizations without quantifying the uncertainty around the criteria.
+While existing software can compute group fairness criteria, they only provide point estimates and/or visualizations without quantifying the uncertainty around the criteria.
 This limitation prevents users from determining whether observed disparities between groups are statistically significant or merely the result of random variation due to finite sample size, potentially leading to incorrect conclusions about fairness violations.
@@
 # Evaluating Fairness Criteria
+
 The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest.

### 2025-06-04 utils/joss_paper/paper.Rmd
adding semicolon

--- 2025-06-04 paper.Rmd
@@
 affiliations:
-  - name: "Department of Statistcial Science, University of Toronto"
+  - name: "Department of Statistical Science, University of Toronto"
     index: 1
@@
 Second, group fairness is the most widely used framework for binary classification tasks [@mehrabi_survey_21].
-Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness [@lum_bias_22].
+Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness [@lum_debias_22].
 This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.
@@
 
-Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022, @Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
+Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022; @Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
 Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.

### 2025-06-04 utils/joss_paper/paper.Rmd
wrapping code output for joss article

@@
 
-#>                  Metric                          Full Metric Name GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#> 1                   PPR                        Statistical Parity        1.12      1.06       0.06    [0.02, 0.1]  1.06 [1.02, 1.09]
-#> 2                   PPR Conditional Statistical Parity (age >=60)        1.23      1.16       0.07      [0, 0.14]  1.06    [1, 1.13]
-#> 3                   FNR                         Equal Opportunity       -0.35     -0.23      -0.12  [-0.26, 0.02]  1.52  [0.9, 2.58]
-#> 4                   FPR                       Predictive Equality        1.07      1.03       0.04   [0.01, 0.07]  1.04 [1.01, 1.07]
-#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.50      0.50       0.00         [0, 0]  1.00       [1, 1]
-#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.50      0.50       0.00         [0, 0]  1.00       [1, 1]
-#> 7                   PPV                Positive Predictive Parity        0.21      0.16       0.05       [0, 0.1]  1.31 [0.99, 1.74]
-#> 8                   NPV                Negative Predictive Parity        0.87      0.88      -0.01  [-0.06, 0.04]  0.99 [0.74, 1.32]
-#> 9           Brier Score                        Brier Score Parity        0.37      0.40      -0.03 [-0.04, -0.02]  0.92 [0.89, 0.96]
-#> 10             Accuracy                   Overall Accuracy Parity        1.00      1.01      -0.01  [-0.05, 0.03]  0.99 [0.95, 1.03]
-#> 11          FN/FP Ratio                        Treatment Equality        0.13      0.12       0.01  [-0.04, 0.06]  1.08  [0.73, 1.6]
+#>    Metric               `Full Metric Name`                        GroupFemale
+#>    <chr>                <chr>                                           <dbl>
+#>  1 PPR                  Statistical Parity                               1.13
+#>  2 PPR                  Conditional Statistical Parity (age >=60)        1.24
+#>  3 FNR                  Equal Opportunity                               -0.38
+#>  4 FPR                  Predictive Equality                              1.07
+#>  5 Avg. Predicted Prob. Balance for Positive Class                       0.5 
+#>  6 Avg. Predicted Prob. Balance for Negative Class                       0.5
+#>  7 PPV                  Positive Predictive Parity                       0.21
+#>  8 NPV                  Negative Predictive Parity                       0.88
+#>  9 Brier Score          Brier Score Parity                               0.37
+#> 10 Accuracy             Overall Accuracy Parity                          1.01
+#> 11 FN/FP Ratio          Treatment Equality                               0.12
+#>    GroupMale Difference `95% Diff CI`  Ratio `95% Ratio CI`
+#>        <dbl>      <dbl> <chr>          <dbl> <chr>
+#>  1      1.06     0.0700 [0.03, 0.11]    1.07 [1.03, 1.1]
+#>  2      1.17     0.0700 [0, 0.14]       1.06 [1, 1.13]
+#>  3     -0.25    -0.130  [-0.27, 0.01]   1.52 [0.96, 2.41]  
+#>  4      1.03     0.0400 [0.01, 0.07]    1.04 [1.01, 1.07]
+#>  5      0.5      0      [0, 0]          1    [1, 1]
+#>  6      0.5      0      [0, 0]          1    [1, 1]
+#>  7      0.17     0.0400 [-0.01, 0.09]   1.24 [0.93, 1.65]
+#>  8      0.89    -0.0100 [-0.07, 0.05]   0.99 [0.74, 1.33]
+#>  9      0.4     -0.0300 [-0.04, -0.02]  0.92 [0.89, 0.96]
+#> 10      1.01     0      [-0.04, 0.04]   1    [0.96, 1.04]  
+#> 11      0.12     0      [-0.05, 0.05]   1    [0.67, 1.49]
+
 ```

### 2025-06-04 R/Fairness.R
typo fix

@@ -1852,7 +1852,17 @@ eval_pos_class_bal <- function(data, outcome, group, probs, confint = TRUE,
       "95% Ratio CI"
     )
 
-  }else{
+
+    if (message) {
+      if (lower_ci > 0 || upper_ci < 0) {
+        cat("There is evidence that the model does not satisfy
+            balance for positive class.\n")
+      } else {
+        cat("There is not enough evidence that the model does not satisfy
+            balance for positive class.\n")
+      }
+    }
+  } else{
     result_df <- data.frame(
       "Avg. Predicted Prob.",
       avg_prob[[1]],
@@ -1868,16 +1878,7 @@ eval_pos_class_bal <- function(data, outcome, group, probs, confint = TRUE,
       "Difference",
       "Ratio"
     )
-  }
 
-  if (message) {
-    if (lower_ci > 0 || upper_ci < 0) {
-      cat("There is evidence that the model does not satisfy
-            balance for positive class.\n")
-    } else {
-      cat("There is not enough evidence that the model does not satisfy
-            balance for positive class.\n")
-    }
   }
 
   return(result_df)

### 2025-06-05 utils/joss_paper/paper.Rmd
fix

--- 2025-06-04 utils/joss_paper/paper.Rmd
@@
 # Make predictions on the test set
-test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")
+test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[,2]
 
@@
 
-#>                  Metric                          Full Metric Name GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#> 1                   PPR                        Statistical Parity        1.12      1.06       0.06    [0.02, 0.1]  1.06 [1.02, 1.09]
-#> 2                   PPR Conditional Statistical Parity (age >=60)        1.23      1.16       0.07      [0, 0.14]  1.06    [1, 1.13]
-#> 3                   FNR                         Equal Opportunity       -0.35     -0.23      -0.12  [-0.26, 0.02]  1.52  [0.9, 2.58]
-#> 4                   FPR                       Predictive Equality        1.07      1.03       0.04   [0.01, 0.07]  1.04 [1.01, 1.07]
-#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.50      0.50       0.00         [0, 0]  1.00       [1, 1]
-#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.50      0.50       0.00         [0, 0]  1.00       [1, 1]
-#> 7                   PPV                Positive Predictive Parity        0.21      0.16       0.05       [0, 0.1]  1.31 [0.99, 1.74]
-#> 8                   NPV                Negative Predictive Parity        0.87      0.88      -0.01  [-0.06, 0.04]  0.99 [0.74, 1.32]
-#> 9           Brier Score                        Brier Score Parity        0.37      0.40      -0.03 [-0.04, -0.02]  0.92 [0.89, 0.96]
-#> 10             Accuracy                   Overall Accuracy Parity        1.00      1.01      -0.01  [-0.05, 0.03]  0.99 [0.95, 1.03]
-#> 11          FN/FP Ratio                        Treatment Equality        0.13      0.12       0.01  [-0.04, 0.06]  1.08  [0.73, 1.6]
+#>                  Metric                          Full Metric Name GroupFemale
+#> 1                   PPR                        Statistical Parity        0.17
+#> 2                   PPR Conditional Statistical Parity (age >=60)        0.34
+#> 3                   FNR                         Equal Opportunity        0.36
+#> 4                   FPR                       Predictive Equality        0.07
+#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.46
+#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.15
+#> 7                   PPV                Positive Predictive Parity        0.64
+#> 8                   NPV                Negative Predictive Parity        0.93
+#> 9           Brier Score                        Brier Score Parity        0.09
+#> 10             Accuracy                   Overall Accuracy Parity        0.88
+#> 11          FN/FP Ratio                        Treatment Equality        1.00
+#>    GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
+#> 1       0.09       0.08   [0.04, 0.12]  1.89 [1.34, 2.67]
+#> 2       0.24       0.10   [0.02, 0.18]  1.42 [1.05, 1.91]
+#> 3       0.58      -0.22 [-0.37, -0.07]  0.62 [0.43, 0.89]
+#> 4       0.03       0.04   [0.01, 0.07]  2.33 [1.19, 4.56]
+#> 5       0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
+#> 6       0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
+#> 7       0.69      -0.05  [-0.21, 0.11]  0.93 [0.72, 1.19]
+#> 8       0.91       0.02  [-0.15, 0.19]  1.02 [0.79, 1.32]
+#> 9       0.08       0.01  [-0.01, 0.03]  1.12 [0.88, 1.43]
+#> 10      0.89      -0.01  [-0.05, 0.03]  0.99 [0.95, 1.03]
+#> 11      3.00      -2.00  [-4.11, 0.11]  0.33 [0.15, 0.73]
+
 ```

### 2025-06-05 utils/joss_paper/paper.Rmd
another fix

@@
 
-#>There is evidence that model does not satisfy equal opportunity.
+#> There is evidence that model does not satisfy equal opportunity.
 #>   Metric GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#>1    FNR        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.85]
+#> 1    FNR        0.36      0.58      -0.22 [-0.37, -0.07]  0.62 [0.43, 0.89]
 ```

### 2025-06-05 utils/joss_paper/paper.Rmd
editing title per JOSS reviewer comments

@@
 ---
-title: '{fairmetrics}: An R package for group fairness evaluation'
+title: 'fairmetrics: An R package for group fairness evaluation'
 tags:

### 2025-06-08 R/get_fairness_metrics.R
Update get_fairness_metrics.R
for better presentation

@@ -98,7 +98,6 @@ get_fairness_metrics <- function(data,
     alpha = alpha,
     digits = digits,
     message = FALSE
-
   )
 
   if (!(is.null(group2) & is.null(condition))) {
@@ -119,22 +118,6 @@ get_fairness_metrics <- function(data,
       digits = digits,
       message = FALSE
     )
-
-    split_condition <- strsplit(condition, split = '')[[1]]
-    if (split_condition[1] == ">" | split_condition[1] == "=") {
-      cond_stats_parity[["Full Metric Name"]] <- paste0("Conditional Statistical Parity (",
-                                              group2,
-                                              " ",
-                                              condition,
-                                              ")")
-    } else {
-      cond_stats_parity[["Full Metric Name"]] <- paste0("Conditional Statistical Parity (",
-                                              condition,
-                                              " ",
-                                              group2,
-                                              ")")
-    }
-
   }
   eq_opp <- eval_eq_opp(
     data = data,
@@ -245,56 +228,48 @@ get_fairness_metrics <- function(data,
     message = FALSE
   )
 
-  # Rename metrics so that its easy to read
-  stats_parity[["Full Metric Name"]] <- "Statistical Parity"
-  eq_opp[["Full Metric Name"]] <- "Equal Opportunity"
-  pred_equality[["Full Metric Name"]] <- "Predictive Equality"
-  pos_class_bal[["Full Metric Name"]] <- "Balance for Positive Class"
-  neg_class_bal[["Full Metric Name"]] <- "Balance for Negative Class"
-  pos_pred_parity[["Full Metric Name"]] <- "Positive Predictive Parity"
-  neg_pred_parity[["Full Metric Name"]] <- "Negative Predictive Parity"
-  bs_parity[["Full Metric Name"]] <- "Brier Score Parity"
-  acc_parity[["Full Metric Name"]] <- "Overall Accuracy Parity"
-  treatment_equality[["Full Metric Name"]] <- "Treatment Equality"
-
+  # Model performance summary
   if (!(is.null(group2) & is.null(condition))) {
-    result <- do.call(
-      rbind,
-      list(
-        stats_parity,
-        cond_stats_parity,
-        eq_opp,
-        pred_equality,
-        pos_class_bal,
-        neg_class_bal,
-        pos_pred_parity,
-        neg_pred_parity,
-        bs_parity,
-        acc_parity,
-        treatment_equality
-      )
-    )
-
[…]

### 2025-06-08 R/Fairness.R
change abb to full name

@@ -123,7 +123,7 @@ eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5,confint = TRUE
 
     # Create a dataframe for the results
     results_df <- data.frame(
-      "FNR",
+      "False Negative Rate",
       fnr[[1]],
       fnr[[2]],
       fnr_diff,
@@ -139,7 +139,7 @@ eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5,confint = TRUE
     )
   } else{
     results_df <- data.frame(
-      "FNR",
+      "False Negative Rate",
       fnr[[1]],
       fnr[[2]],
       fnr_diff,
@@ -471,7 +471,7 @@ eval_stats_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint
 
     # Structure the results as a dataframe
     results_df <- data.frame(
-      Metric = "PPR",
+      Metric = "Positive Prediction Rate",
       Group1 = ppr[[1]],
       Group2 = ppr[[2]],
       Difference = ppr_diff,
@@ -489,7 +489,7 @@ eval_stats_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint
   }else{
     # Structure the results as a dataframe
     results_df <- data.frame(
-      Metric = "PPR",
+      Metric = "Positive Prediction Rate",
       Group1 = ppr[[1]],
       Group2 = ppr[[2]],
       Difference = ppr_diff,
@@ -747,7 +747,7 @@ eval_pos_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, conf
     upper_ratio_ci <- round(exp(log(ppv_ratio) + qnorm(1 - alpha / 2) * sd(se[2, ])), digits)
 
     result_df <- data.frame(
-      "PPV",
+      "Positive Predictive Value",
       ppv[[1]],
       ppv[[2]],
       ppv_dif,
@@ -766,7 +766,7 @@ eval_pos_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, conf
     )
   }else{
     result_df <- data.frame(
-      "PPV",
+      "Positive Predictive Value",
       ppv[[1]],
       ppv[[2]],
       ppv_dif,
@@ -899,7 +899,7 @@ eval_neg_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, conf
[…]

### 2025-07-21 utils/joss_paper/paper.Rmd
removing sentence, keeping citation

@@
     index: 1
-date: "2025-06-02"
+date: "2025-06-10"
 bibliography: paper.bib
 output:
-  rticles::joss_article
-  #rticles::arxiv_article
+  rticles::joss_article:
+    keep_tex: true
   # md_document:
+  #    keep_tex: true
   #    preserve_yaml: TRUE
@@
 
-Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for specific groups, particularly those defined by protected attributes such as race, gender, or age.
-Evaluating fairness is a critical aspect of ML model development, as biased models can perpetuate structural inequalities.
-The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity).
-Group-based fairness criteria assess whether a model is equally accurate or well-calibrated across a set of predefined groups so that appropriate bias mitigation strategies can be implemented.
-{fairmetrics} provides both point and interval estimates for multiple metrics through a convenient wrapper function and includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].
+Fairness is a growing area of machine learning (ML) that focuses on ensuring that models do not produce systematically biased outcomes across groups defined by protected attributes, such as race, gender, or age.
+The `fairmetrics` R package provides a user-friendly framework for rigorously evaluating group-based fairness criteria, including independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity).
+The package provides both point and interval estimates for a variety of commonly used criteria.
+`fairmetrics` also includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical] to demonstrate its use.
 
@@
 
-ML models are increasingly integrated into high-stakes domains to support decision making that significantly impacts individuals and society more broadly, including criminal justice, healthcare, finance, employment, and education [@mehrabi_survey_21].
-Mounting evidence suggest that these models often exhibit bias across groups defined by protected attributes.
-For example, within criminal justice, the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) software, a tool used by U.S.
-courts to evaluate the risk of defendants becoming recidivists, was found to incorrectly classify Black defendants as high-risk at nearly twice the rate of white defendants [@mattuMachineBias].
-This bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to white defendants with similar risk profiles.
-Similarly, within healthcare, a commercial risk-prediction algorithm deployed in the U.S.
-to identify patients with complex health needs for high-risk care management programs was shown to be significantly less calibrated for Black patients relative to white patients [@obermeyerDissectingRacialBias2019].
-This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to white patients.
-These examples illustrate that there is an urgent need for practitioners and researchers to ensure that ML models support fair decision making before they are deployed in real-world applications.
-
-While existing software can compute group fairness criteria, they only provide point estimates and/or visualizations without quantifying the uncertainty around the criteria.
-This limitation prevents users from determining whether observed disparities between groups are statistically significant or merely the result of random variation due to finite sample size, potentially leading to incorrect conclusions about fairness violations.
-The {fairmetrics} R package addresses this gap by providing bootstrap-based confidence intervals (CIs) for both difference-based and ratio-based group fairness metrics, empowering users to make statistically grounded decisions about the fairness of their models, which is inconsistently done in practice.
-
-# Scope
-
-The `{fairmetrics}` package is designed to evaluate group fairness in the setting of binary classification with a binary protected attribute.
-This restriction reflects standard practice in the fairness literature and is motivated by several considerations.
-First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative [@mehrabi_survey_21].
-Second, group fairness is the most widely used framework for binary classification tasks [@mehrabi_survey_21].
-Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness [@lum_debias_22].
-This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.
+ML models are increasingly used in high-stakes domains such as criminal justice, healthcare, finance, employment, and education [@mehrabi_survey_21; @mattuMachineBias; @Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
+Existing fairness evaluation software report point estimates and/or visualizations, without any measures of uncertainty.
+This limits users' ability to determine whether observed disparities are statistically significant.
+`fairmetrics` addresses this limitation by including confidence intervals for both difference and ratio based fairness metrics to enable more robust and statistically grounded fairness assessments.
 
@@
 
-Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022; @Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024].
-Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups.
-Separation requires independence between the classifications and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within the positive (or negative) outcome class.
-Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model's prediction is known, the protected attribute provides no additional information about the true outcome.
-Below we summarize the fairness metrics that are available within the {fairmetrics} package.
+`fairmetrics` is designed to evaluate fairness of binary classification models across binary protected attributes.
+The package supports the evaluation of metrics belonging to three major group fairness criteria:
 
-## Independence
+- **Independence:** Statistical Parity (compares the overall rate of positive predictions between groups), Conditional Statistical Parity (compares the rate of positive predictions between groups within a specific subgroup).
 
--   **Statistical Parity:** Compares the overall rate of positive predictions between groups.
+- **Separation:** Equal Opportunity (compares false negative rates between groups), Predictive Equality (compares false positive rates between groups), Balance for Positive Class (compares the average predicted probabilities among individuals whose true outcome is positive across groups), and Balance for Negative Class (compares the average predicted probabilities among individuals whose true outcome is negative across groups).
 
--   **Conditional Statistical Parity:** Restricts the comparison of positive prediction rates to a specific subgroup (e.g., within a hospital unit or age bracket), offering a more context-specific fairness assessment.
+- **Sufficiency:** Positive Predictive Parity (compares the positive predictive values across groups), Negative Predictive Parity (compares the negative predictive values across groups).
 
-## Separation
-
--   **Equal Opportunity:**  Compares disparities in the false negative rates between groups, quantifying differences in the likelihood of missing positive outcomes.
-
--   **Predictive Equality:** Compares the false positive rates (FPR) between groups, quantifying differences in the likelihood of incorrectly labeling negative outcomes as positive.
-
--   **Balance for Positive Class:**  Compares the average of the predicted probabilities among individuals whose true outcome is positive across groups.
-
--   **Balance for Negative Class:** Compares the average of the predicted probabilities among individuals whose true outcome is negative across groups.
-
-## Sufficiency
-
--   **Positive Predictive Parity:** Compares the positive predictive values across groups, assessing differences in the precision of positive predictions.
-
--   **Negative Predictive Parity:** Compares the negative predictive values across groups, assessing differences in the precision of negative predictions.
-
-## Other Criteria
-
--   **Brier Score Parity:** Compares the Brier score (i.e., the mean squared error of the predicted probabilities) across groups, evaluating differences in calibration.
-
--   **Accuracy Parity:** Compares the overall accuracy of a predictive model across groups.
-
--   **Treatment Equality:** Compares the ratio of false negatives to false positives across groups, evaluating whether the trade-off between missed detections of positive outcomes and false alarms of negative outcomes is balanced.
+The package also includes additional metrics, such as the Brier Score Parity (compares the Brier score across groups), Accuracy Parity (compares the overall accuracy across groups), and Treatment Equality (compares the ratio of false negatives to false positives across groups).
 
@@
 
-The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest.
-\hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}.
-Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function.
+The input required to evaluate model fairness with the `fairmetrics` package is a `data.frame` or `tibble` containing the model's predicted probabilities, the true outcomes, and the protected attribute.
+\hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using `fairmetrics`.
 
-![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow.png)
+![Workflow for using `fairmetrics` to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow.png)
 
-A simple example of how to use the {fairmetrics} package is illustrated below.
-The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from the MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in the {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data].
-The dataset consists of 1,776 hemodynamically stable patients with respiratory failure and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating whether the patient died within 28 days of hospital admission.
+A simple example of how to use the `fairmetrics` package is illustrated below.
+The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the Indwelling Arterial Catheter (IAC) Clinical dataset, from the MIMIC-II clinical database [@goldberger2000physiobank; @raffa2016clinical; @raffa2016data].
 
-While the choice of fairness criteria used is context dependent, we show all criteria available with the `get_fairness_metrics()` function for the purposes of illustration.
+While the choice of fairness metric used is context dependent, we show all criteria available with the `get_fairness_metrics()` function for illustrative purposes.
 In this example, we evaluate the model's fairness with respect to the protected attribute `gender`.
-For conditional statistical parity, we condition on patients older than 60 years old.
+For conditional statistical parity, we condition on age greater than 60 years.
 The model is trained on a subset of the data and the predictions are made and evaluated on a test set.
-The `get_fairness_metrics()` function outputs difference and ratio-based metrics as well as their corresponding confidence intervals.
 A statistically significant difference across groups at a given level of significance is indicated when the confidence interval for a difference-based metric does not include zero or when the interval for a ratio-based metric does not include one.
@@
 ```r
+# Train a classification model (e.g., random forest).
+# Add the vector of predicted probabilities to the test data
+# to evaluate fairness.
 library(fairmetrics)
-library(dplyr)
-library(magrittr)
-library(randomForest)
-
-# Load the example dataset
-data("mimic_preprocessed")
-
-# Split the data into training and test sets
-train_data <- mimic_preprocessed %>%
-  dplyr::filter(dplyr::row_number() <= 700)
-
-test_data <- mimic_preprocessed %>%
-  dplyr::mutate(gender = ifelse(gender_num == 1, "Male", "Female")) %>%
-  dplyr::filter(dplyr::row_number() > 700)
-
-# Train a random forest model
-rf_model <- randomForest::randomForest(
-  factor(day_28_flg) ~ ., 
-  data = train_data, 
-  ntree = 1000
-  )
-
-# Make predictions on the test set
-test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[,2]
-
-# Get fairness metrics
 # Setting alpha=0.05 for 95% confidence intervals
@@
 
-#>                  Metric                          Full Metric Name GroupFemale
-#> 1                   PPR                        Statistical Parity        0.17
-#> 2                   PPR Conditional Statistical Parity (age >=60)        0.34
-#> 3                   FNR                         Equal Opportunity        0.36
-#> 4                   FPR                       Predictive Equality        0.07
-#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.46
-#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.15
-#> 7                   PPV                Positive Predictive Parity        0.64
-#> 8                   NPV                Negative Predictive Parity        0.93
-#> 9           Brier Score                        Brier Score Parity        0.09
-#> 10             Accuracy                   Overall Accuracy Parity        0.88
-#> 11          FN/FP Ratio                        Treatment Equality        1.00
-#>    GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#> 1       0.09       0.08   [0.04, 0.12]  1.89 [1.34, 2.67]
-#> 2       0.24       0.10   [0.02, 0.18]  1.42 [1.05, 1.91]
-#> 3       0.58      -0.22 [-0.37, -0.07]  0.62 [0.43, 0.89]
-#> 4       0.03       0.04   [0.01, 0.07]  2.33 [1.19, 4.56]
-#> 5       0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
-#> 6       0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
-#> 7       0.69      -0.05  [-0.21, 0.11]  0.93 [0.72, 1.19]
-#> 8       0.91       0.02  [-0.15, 0.19]  1.02 [0.79, 1.32]
-#> 9       0.08       0.01  [-0.01, 0.03]  1.12 [0.88, 1.43]
-#> 10      0.89      -0.01  [-0.05, 0.03]  0.99 [0.95, 1.03]
-#> 11      3.00      -2.00  [-4.11, 0.11]  0.33 [0.15, 0.73]
+$performance
+                                    Metric GroupFemale GroupMale
+1                 Positive Prediction Rate        0.18      0.09
+2                 Positive Prediction Rate        0.35      0.24
+3                      False Negative Rate        0.36      0.59
+4                      False Positive Rate        0.08      0.03
+5                     Avg. Predicted Prob.        0.47      0.37
+6                     Avg. Predicted Prob.        0.16      0.11
+7                Positive Predictive Value        0.62      0.67
+8                Negative Predictive Value        0.93      0.91
+9                              Brier Score        0.09      0.08
+10                                Accuracy        0.87      0.89
+11 (False Negative)/(False Positive) Ratio        0.93      2.89
 
+$fairness
+                           Metric Difference    95% Diff CI Ratio 95% Ratio CI
+1              Statistical Parity       0.08   [0.04, 0.12]  2.00 [1.41, 2.83]
+2  Conditional Statistical Parity       0.10   [0.02, 0.18]  1.43 [1.05, 1.96]
+3               Equal Opportunity      -0.21 [-0.36, -0.06]  0.64  [0.45, 0.9]
+4             Predictive Equality       0.04   [0.01, 0.07]  2.33  [1.13, 4.8]
+5      Balance for Positive Class       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
+6      Balance for Negative Class       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
+7      Positive Predictive Parity      -0.06  [-0.23, 0.11]  0.92 [0.71, 1.18]
+8      Negative Predictive Parity       0.01  [-0.15, 0.17]  1.01 [0.79, 1.29]
+9              Brier Score Parity       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
+10        Overall Accuracy Parity      -0.01  [-0.05, 0.03]  0.99 [0.95, 1.03]
+11             Treatment Equality      -2.28  [-4.72, 0.16]  0.33 [0.15, 0.73]
 ```
 
-Should the user wish to calculate an individual criteria, it is possible to use any of the `eval_*` functions.
-For example, to calculate equal opportunity, the user can call the `eval_equal_opportunity()` function.
-
-```r
-eval_eq_opp(
-  data = test_data,
-  outcome = "day_28_flg",
-  group = "gender",
-  probs = "pred",
-  confint = TRUE,
-  cutoff = 0.41,
-  alpha = 0.05
-)
-
-#> There is evidence that model does not satisfy equal opportunity.
-#>   Metric GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
-#> 1    FNR        0.36      0.58      -0.22 [-0.37, -0.07]  0.62 [0.43, 0.89]
-```
+Users can also compute individual metrics using functions like `eval_eq_opp()` to 
+test specific fairness conditions. Full usage examples are provided in the package documentation.
 
@@
 
-Other R packages similar to {fairmetrics} include {fairness} [@fairness_package], {fairmodels} [@wisniewski2022fairmodels] and {mlr3fairness} [@mlr3fairness_package].
-The differences between {fairmetrics} and these other packages is twofold.
-The primary difference between is that {fairmetrics} calculates the ratio and difference between group fairness criterion and their corresponding confidence intervals of fairness metrics via bootstrap, allowing for more meaningful inferences about the fairness criteria.
-Additionally, in contrast to the {fairmodels}, {fairness} and {mlr3fairness} packages, the {fairmetrics} package does not posses any external dependencies and has a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems.
+Other R packages similar to `fairmetrics` include `fairness` [@fairness_package], `fairmodels` [@wisniewski2022fairmodels] and `mlr3fairness` [@mlr3fairness_package].
+`fairmetrics` differs from these packages in two ways.
+The first difference is that `fairmetrics` calculates ratio and difference-based group fairness metrics and their corresponding confidence intervals, allowing for more meaningful inferences about the fairness criteria.
+The second difference is that `fairmetrics` does not possess any external dependencies and has a lower memory footprint.
 \hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows the comparison of memory used and dependencies required when loading each library.
@@
 fairness    & 117.61 & 141\\
-fairmodels  & 58.11  & 45 \\
+mlr3fairness & 58.11  & 45 \\
 fairmetrics & 0.05   & 0  \\
@@
 \end{tabular}
-\caption{Memory usage (in MB) and dependencies of {fairmetrics} vs similar packages.}
+\caption{Memory usage (in MB) and dependencies of `fairmetrics` vs similar packages.}
 \label{tab:memory_dep_usage}
@@
 
-For python users, the {fairlearn} library [@fairlearn_paper] provides additional fairness metrics and algorithms.
-The {fairmetrics} package is designed for seemless integration with R workflows, making it a more convenient choice for R-based ML applications.
+For Python users, the `fairlearn` library [@fairlearn_paper] provides additional fairness metrics and algorithms.
+The `fairmetrics` package is designed for seamless integration with R workflows, making it a more convenient choice for R users.
 
@@
 
-The {fairmetrics} package is under the MIT license.
+The `fairmetrics` package is under the MIT license.
 It is available on CRAN and can be installed by using `install.packages("fairmetrics")`.
-A more in-depth tutorial can be accessed at: https://jianhuig.github.io/fairmetrics/articles/fairmetrics.html.
-All code is open-source and hosted on
-GitHub. All bugs and inquiries can be reported at https://github.com/jianhuig/fairmetrics/issues/.
+Full documentation and its examples are available at: https://jianhuig.github.io/fairmetrics/articles/fairmetrics.html.
+Source code and issue tracking are hosted on GitHub: https://github.com/jianhuig/fairmetrics/.
 

### 2025-07-21 R/Fairness.R
adding new tests

@@ -78,7 +78,7 @@
 #' }
 #' @export
 
-eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5,confint = TRUE,
+eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                         bootstraps = 2500, alpha = 0.05, digits = 2,
                         message = TRUE) {
   # Check if outcome is binary

### 2025-07-21 tests/testthat/test-Fairness-outcomes.R

@@ -1,0 +1,247 @@
+test_that('eval_acc_parity ratio+diff outcome test', {
+  expect_equal({
+    dt <- data.frame(
+      y_true = c(0, 1, 0, 1),
+      y_pred = c(1, 1, 1, 1),
+      group = c("A", "A", "B", "B")
+    )
+    result <- eval_acc_parity(data = dt,
+                              outcome = "y_true",
+                              group = "group",
+                              probs = "y_pred",
+                              digits = 1,
+                              confint = TRUE,
+                              message = FALSE)
+    list(result$Ratio, result$Difference)
+  },
+  list(1,0))
+})
+
+test_that('eval_acc_parity CI outcome test', {
+  expect_equal({
+    dt <- data.frame(
+      y_true = c(0, 1, 0, 1),
+      y_pred = c(1,1,1,1),
+      group = c("A", "A", "B", "B")
+    )
+    result <- eval_acc_parity(data = dt,
+                              outcome = "y_true",
+                              group = "group",
+                              probs = "y_pred",
+                              digits = 1,
+                              confint = TRUE,
+                              message = FALSE)
+    list(result$`95% Ratio CI`, result$`95% Diff CI`)
+  },
+  list("[NA, NA]", "[-1, 1]"))
+})
+
+
+test_that('eval_bs_parity ratio+diff outcome test', {
+  expect_equal({
+    dt <- data.frame(
+      y_true = c(0, 1, 0, 1),
+      y_pred = c(1, 1, 1, 1),
+      group = c("A", "A", "B", "B")
+    )
+    result <- eval_bs_parity(data = dt,
+                              outcome = "y_true",
+                              group = "group",
+                              probs = "y_pred",
+                              digits = 1,
+                              confint = TRUE,
+                              message = FALSE)
+    list(result$Ratio, result$Difference)
+  },
+  list(1,0))
+})
+
+test_that('eval_bs_parity CI outcome test', {
+  expect_equal({
+    dt <- data.frame(
+      y_true = c(0, 1, 0, 1),
+      y_pred = c(1,1,1,1),
+      group = c("A", "A", "B", "B")
+    )
+    result <- eval_bs_parity(data = dt,
+                              outcome = "y_true",
+                              group = "group",
+                              probs = "y_pred",
+                              digits = 1,
+                              confint = TRUE,
+                              message = FALSE)
[…]

### 2025-08-04 utils/joss_paper/paper.Rmd
edited README, description, and paper

@@
     affiliation: 1
+  - name: Benson Chou
+    orcid: 0009-0007-0265-033X
+    affiliation: 1
   - name: Jessica Gronsbell
@@
 output:
-  rticles::joss_article:
-    keep_tex: true
-  # md_document:
+  # rticles::joss_article:
   #   keep_tex: true
-  #    preserve_yaml: TRUE
-  #    variant: "markdown_strict"
+  md_document:
+     #keep_tex: true
+     preserve_yaml: TRUE
+     variant: "markdown_strict"
 journal: JOSS
@@
 Fairness is a growing area of machine learning (ML) that focuses on ensuring that models do not produce systematically biased outcomes across groups defined by protected attributes, such as race, gender, or age.
-The `fairmetrics` R package provides a user-friendly framework for rigorously evaluating group-based fairness criteria, including independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity).
+The `fairmetrics` R package provides a user-friendly framework for rigorously evaluating group-based fairness criteria, including independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity) for binary protected attributes.
 The package provides both point and interval estimates for a variety of commonly used criteria.
@@
 While the choice of fairness metric used is context dependent, we show all criteria available with the `get_fairness_metrics()` function for illustrative purposes.
-In this example, we evaluate the model's fairness with respect to the protected attribute `gender`.
+In this example, we evaluate the model's fairness with respect to the binary protected attribute `gender`.
 For conditional statistical parity, we condition on age greater than 60 years.
@@
 $performance
-                                    Metric GroupFemale GroupMale
-1                 Positive Prediction Rate        0.18      0.09
-2                 Positive Prediction Rate        0.35      0.24
-3                      False Negative Rate        0.36      0.59
+                                    Metric Group1 Group2
+1                 Positive Prediction Rate   0.17   0.08
+2     Conditional Positive Prediction Rate   0.34   0.21
+3                      False Negative Rate   0.38   0.62
 4                      False Positive Rate   0.08   0.03
-5                     Avg. Predicted Prob.        0.47      0.37
-6                     Avg. Predicted Prob.        0.16      0.11
-7                Positive Predictive Value        0.62      0.67
-8                Negative Predictive Value        0.93      0.91
+5            Avg. Predicted Positive Prob.   0.46   0.37
+6            Avg. Predicted Negative Prob.   0.15   0.10
+7                Positive Predictive Value   0.62   0.66
+8                Negative Predictive Value   0.92   0.90
 9                              Brier Score   0.09   0.08
-10                                Accuracy        0.87      0.89
-11 (False Negative)/(False Positive) Ratio        0.93      2.89
+10                                Accuracy   0.87   0.88
+11 (False Negative)/(False Positive) Ratio   1.03   3.24
 
 $fairness
-                           Metric Difference    95% Diff CI Ratio 95% Ratio CI
-1              Statistical Parity       0.08   [0.04, 0.12]  2.00 [1.41, 2.83]
-2  Conditional Statistical Parity       0.10   [0.02, 0.18]  1.43 [1.05, 1.96]
-3               Equal Opportunity      -0.21 [-0.36, -0.06]  0.64  [0.45, 0.9]
-4             Predictive Equality       0.04   [0.01, 0.07]  2.33  [1.13, 4.8]
+                           Metric Difference        Diff_CI Ratio     Ratio_CI
+1              Statistical Parity       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
+2  Conditional Statistical Parity       0.13   [0.05, 0.21]  1.62 [1.17, 2.24]
+3               Equal Opportunity      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
+4             Predictive Equality       0.05   [0.02, 0.08]  2.67 [1.38, 5.15]
 5      Balance for Positive Class       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
 6      Balance for Negative Class       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
-7      Positive Predictive Parity      -0.06  [-0.23, 0.11]  0.92 [0.71, 1.18]
-8      Negative Predictive Parity       0.01  [-0.15, 0.17]  1.01 [0.79, 1.29]
+7      Positive Predictive Parity      -0.04  [-0.21, 0.13]  0.94 [0.71, 1.24]
+8      Negative Predictive Parity       0.02  [-0.02, 0.06]  1.02 [0.98, 1.07]
 9              Brier Score Parity       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
-10        Overall Accuracy Parity      -0.01  [-0.05, 0.03]  0.99 [0.95, 1.03]
-11             Treatment Equality      -2.28  [-4.72, 0.16]  0.33 [0.15, 0.73]
+10        Overall Accuracy Parity      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
+11             Treatment Equality      -2.21  [-4.44, 0.02]  0.32  [0.14, 0.7]
 ```
 
-Users can also compute individual metrics using functions like `eval_eq_opp()` to 
-test specific fairness conditions. Full usage examples are provided in the package documentation.
+Users can also compute individual metrics using functions like `eval_eq_opp()` to test specific fairness conditions. Full usage examples are provided in the package documentation.
 

### 2025-08-05 R/ModelPerformance.R
redocumenting

@@ -4,9 +4,9 @@
 
 #' Calculate the true positive rate
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' protected attribute
 #' @param outcome the name of the outcome variable, it must be binary
-#' @param group the name of the sensitive attribute
+#' @param group the name of the protected attribute
 #' @param probs the name of the predicted outcome variable
 #' @param cutoff the threshold for the predicted outcome, default is 0.5
 #' @param digits the number of digits to round the result to, default is 2
@@ -29,9 +29,9 @@ get_tpr <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
 
 #' Calculate the false positive rate
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' protected attribute
 #' @param outcome the name of the outcome variable, it must be binary
-#' @param group the name of the sensitive attribute
+#' @param group the name of the protected attribute
 #' @param probs the name of the predicted outcome variable
 #' @param cutoff the threshold for the predicted outcome, default is 0.5
 #' @param digits the number of digits to round the result to, default is 2
@@ -53,9 +53,9 @@ get_fpr <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
 
 #' Calculate the probability of positive prediction
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' protected attribute
 #' @param outcome the name of the outcome variable, it must be binary
-#' @param group the name of the sensitive attribute
+#' @param group the name of the protected attribute
 #' @param probs the name of the predicted outcome variable
 #' @param cutoff the threshold for the predicted outcome, default is 0.5
 #' @param digits the number of digits to round the result to, default is 2
@@ -78,9 +78,9 @@ get_ppr <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
 
 #' Calculate the positive predictive value
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' protected attribute
 #' @param outcome the name of the outcome variable, it must be binary
-#' @param group the name of the sensitive attribute
+#' @param group the name of the protected attribute
 #' @param probs the name of the predicted outcome variable
 #' @param cutoff the threshold for the predicted outcome, default is 0.5
 #' @param digits the number of digits to round the result to, default is 2
[…]

### 2025-08-05 R/eval_acc_parity.R

@@ -1,12 +1,11 @@
 #' Examine Accuracy Parity of a Model
 #'
-#' This function assesses *Accuracy Parity*, a fairness criterion that evaluates whether
-#' the overall accuracy of a predictive model is consistent across different groups.
+#' This function assesses *Accuracy Parity*, a fairness criterion that evaluates whether the overall accuracy of a predictive model is consistent across two groups defined by a binary protected attribute.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable
-#' @param group Name of the sensitive attribute
+#' @param group Name of the binary protected attribute. Must consist of only two groups.
 #' @param probs Predicted probabilities
 #' @param cutoff Cutoff value for the predicted probabilities
 #' @param confint Logical indicating whether to calculate confidence intervals
@@ -46,7 +45,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Accuracy Parity
@@ -65,10 +64,14 @@
 eval_acc_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                             alpha = 0.05, bootstraps = 2500,
                             digits = 2, message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
+  }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
   }
 
   acc <- get_acc(

### 2025-08-05 R/eval_bs_parity.R

@@ -1,13 +1,11 @@
 #' Examine Brier Score Parity of a Model
 #'
-#' This function evaluates *Brier Score Parity*, a fairness measure that checks whether the Brier score
-#' (a measure of the calibration of probabilistic predictions) is similar across different groups. Brier score
-#' parity ensures that the model's predicted probabilities are equally well calibrated across subpopulations.
+#' This function evaluates *Brier Score Parity*, a fairness measure that checks whether the Brier score (a measure of the calibration of probabilistic predictions) is similar across across two groups defined by a binary protected attribute. Brier score parity ensures that the model's predicted probabilities are equally well calibrated across subpopulations.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable
-#' @param group Name of the sensitive attribute
+#' @param group Name of the binary protected attribute. Must consist of only two groups.
 #' @param probs Predicted probabilities
 #' @param bootstraps Number of bootstraps to use for confidence intervals
 #' @param confint Logical indicating whether to calculate confidence intervals
@@ -46,7 +44,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #'
 #' # Evaluate Brier Score Parity
 #' eval_bs_parity(
@@ -63,12 +61,15 @@
 eval_bs_parity <- function(data, outcome, group, probs, confint = TRUE,
                            alpha = 0.05, bootstraps = 2500,
                            digits = 2, message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
   }
-
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
+  }
   bs <- get_brier_score(
     data = data, outcome = outcome, group = group, probs = probs,
     digits = digits

### 2025-08-05 R/eval_cond_acc_equality.R

@@ -1,13 +1,11 @@
 #' Examine Conditional Use Accuracy Equality of a Model
 #'
-#' This function evaluates *Conditional Use Accuracy Equality*, a fairness criterion
-#' that requires predictive performance to be similar across groups when a model
-#' makes positive or negative predictions.
+#' This function evaluates *Conditional Use Accuracy Equality*, a fairness criterion that requires predictive performance to be similar across across two groups - defined by a binary protected attribute - when a model makes positive or negative predictions.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable, it must be binary
-#' @param group Name of the sensitive attribute
+#' @param group Name of the binary protected attribute. Must consist of only two groups.
 #' @param probs Name of the predicted outcome variable
 #' @param cutoff Threshold for the predicted outcome, default is 0.5
 #' @param confint Whether to compute 95% confidence interval, default is TRUE
@@ -51,7 +49,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Conditional Use Accuracy Equality
@@ -69,18 +67,15 @@
 eval_cond_acc_equality <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                                    alpha = 0.05, bootstraps = 2500,
                                    digits = 2, message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
   }
-
-  # Check if outcome is binary
-  unique_values <- unique(data[[outcome]])
-  if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+  if (!(length(groups) == 2)) {
[…]

### 2025-08-05 R/eval_cond_stats_parity.R

@@ -1,14 +1,14 @@
 #' Examine Conditional Statistical Parity of a Model
 #'
-#' This function evaluates *conditional statistical parity*, which measures fairness by comparing positive prediction rates across sensitive groups within a defined subgroup of the population. This is useful in scenarios where fairness should be evaluated in a more context-specific way—e.g., within a particular hospital unit or age bracket. Conditional statistical parity is a refinement of standard statistical parity. Instead of comparing prediction rates across groups in the entire dataset, it restricts the comparison
+#' This function evaluates *conditional statistical parity*, which measures fairness by comparing positive prediction rates across two groups defined by a binary protected attribute within a defined subgroup of the population. This is useful in scenarios where fairness should be evaluated in a more context-specific way—e.g., within a particular hospital unit or age bracket. Conditional statistical parity is a refinement of standard statistical parity. Instead of comparing prediction rates across groups in the entire dataset, it restricts the comparison
 #' to a specified subset of the population, defined by a conditioning variable.
 #'
 #' The function supports both categorical and continuous conditioning variables. For continuous variables, you can supply a threshold expression like `"<50"` or `">=75"` to the \code{condition} parameter.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable, it must be binary
-#' @param group Name of the sensitive attribute
+#' @param group Name of the binary protected attribute. Must consist of only two groups.
 #' @param group2 Name of the group to condition on
 #' @param condition If the conditional group is categorical, the condition
 #' supplied must be a character of the levels to condition on. If the conditional
@@ -57,7 +57,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Conditional Statistical Parity
@@ -82,10 +82,14 @@ eval_cond_stats_parity <- function(data, outcome, group,
                                    bootstraps = 2500, alpha = 0.05,
                                    message = TRUE,
                                    digits = 2) {
[…]

### 2025-08-05 R/eval_eq_odds.R

@@ -2,17 +2,16 @@
 #'
 #' This function evaluates whether a predictive model satisfies the Equalized
 #' Odds criterion by comparing both False Negative Rates (FNR) and False
-#' Positive Rates (FPR) across two groups defined by a binary sensitive
+#' Positive Rates (FPR) across two groups defined by a binary protected
 #' attribute. It reports the rate for each group, their differences, ratios, and
 #' bootstrap-based confidence regions. A Bonferroni-corrected union test is used
 #' to test whether the model violates the Equalized Odds criterion.
 #'
 #' @param data A data frame containing the true binary outcomes, predicted
-#' probabilities, and sensitive group membership.
+#' probabilities, and binary protected attribute.
 #' @param outcome A string specifying the name of the binary outcome variable in
 #' \code{data}.
-#' @param group A string specifying the name of the binary sensitive attribute
-#' variable (e.g., race, gender) used to define the comparison groups.
+#' @param group Name of the binary protected attribute. Must consist of only two groups.
 #' @param probs A string specifying the name of the variable containing
 #' predicted probabilities or risk scores.
 #' @param cutoff A numeric value used to threshold predicted probabilities into
@@ -64,7 +63,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Equalized Odds
@@ -80,10 +79,14 @@
 eval_eq_odds <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                          bootstraps = 2500, alpha = 0.05, digits = 2,
                          message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
+  }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
   }
 
   fnr <- 1 - get_tpr(

### 2025-08-05 R/eval_eq_opp.R

@@ -2,17 +2,16 @@
 #'
 #' This function evaluates the fairness of a predictive model with respect to
 #' the Equal Opportunity criterion, which requires that the False Negative Rate
-#' (FNR) be comparable across groups defined by a sensitive attribute. The
+#' (FNR) be comparable across groups defined by a binary protected attribute. The
 #' function quantifies disparities in FNR between two groups and provides both
 #' the absolute difference and ratio, along with confidence intervals obtained
 #' via bootstrapping.
 #'
 #' @param data A data frame containing the true binary outcomes, predicted
-#' probabilities, and sensitive group membership.
+#' probabilities, and binary protected attribute.
 #' @param outcome A string specifying the name of the binary outcome variable in
 #' \code{data}.
-#' @param group A string specifying the name of the sensitive attribute variable
-#' (e.g., race, gender).
+#' @param group group Name of the binary protected attribute. Must consist of only two groups.
 #' @param probs A string specifying the name of the variable containing
 #' predicted probabilities or risk scores.
 #' @param cutoff A numeric value used to threshold predicted probabilities into
@@ -64,7 +63,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Equal Opportunity Compliance
@@ -83,10 +82,15 @@ eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRU
                         message = TRUE) {
   # Check if outcome is binary
   unique_values <- unique(data[[outcome]])
+  # Check if outcome and groups are binary
+  unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
   }
-
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
+  }
   fnr <- 1 - get_tpr(
     data = data, outcome = outcome, group = group, probs = probs,
     cutoff = cutoff, digits = digits

### 2025-08-05 R/eval_neg_class_bal.R

@@ -1,13 +1,12 @@
 #' Examine Balance for Negative Class of a Model
 #'
 #' This function evaluates *Balance for the Negative Class*, a fairness criterion
-#' that checks whether the model assigns similar predicted probabilities across groups
-#' among individuals whose true outcome is negative (i.e. \eqn{Y = 0}).
+#' that checks whether the model assigns similar predicted probabilities among individuals whose true outcome is negative (i.e. \eqn{Y = 0}) accross groups defined by a binary protected attribute.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute attribute
 #' @param outcome Name of the outcome variable
-#' @param group Name of the sensitive attribute
+#' @param group Name of the protected attribute. Must consist of only two groups.
 #' @param probs Predicted probabilities
 #' @param confint Logical indicating whether to calculate confidence intervals
 #' @param bootstraps Number of bootstraps to use for confidence intervals
@@ -46,7 +45,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #'
 #' # Evaluate Balance for Negative Class
 #' eval_neg_class_bal(
@@ -62,10 +61,14 @@
 eval_neg_class_bal <- function(data, outcome, group, probs, confint = TRUE,
                                alpha = 0.05, bootstraps = 2500,
                                digits = 2, message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
+  }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
   }
 
   neg_data <- data[data[[outcome]] == 0, ]

### 2025-08-05 R/eval_neg_pred_parity.R

@@ -1,14 +1,14 @@
 #' Examine Negative Predictive Parity of a Model
 #'
 #' This function evaluates *negative predictive predictive parity*, a key fairness criterion that
-#' compares the *Negative Predictive Value (NPV)* between groups defined by a sensitive attribute.
+#' compares the *Negative Predictive Value (NPV)* between groups defined by a binary protected attribute.
 #' In other words, it assesses whether, among individuals predicted to be negative,
 #' the probability of being truly negative is equal across subgroups.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' protected attribute
 #' @param outcome Name of the outcome variable, it must be binary
-#' @param group Name of the sensitive attribute
+#' @param group Name of the protected attribute. Must consist of only two groups.
 #' @param probs Name of the predicted outcome variable
 #' @param cutoff Threshold for the predicted outcome, default is 0.5
 #' @param confint Whether to compute 95% confidence interval, default is TRUE
@@ -50,7 +50,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Negative Predictive Parity
@@ -68,10 +68,14 @@
 eval_neg_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                                  bootstraps = 2500, alpha = 0.05,
                                  digits = 2, message = TRUE) {
-  # Check if outcome is binary
-  unique_values <- sort(unique(data[[outcome]]))
+  # Check if outcome and groups are binary
+  unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
+  }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
   }
 
   npv <- get_npv(

### 2025-08-05 R/eval_pos_class_bal.R

@@ -1,13 +1,12 @@
 #' Examine Balance for the Positive Class of a Model
 #'
 #' This function evaluates *Balance for the Positive Class*, a fairness criterion
-#' that checks whether the model assigns similar predicted probabilities across groups
-#' among individuals whose true outcome is positive (i.e. \eqn{Y = 1}).
+#' that checks whether the model assigns similar predicted probabilities among individuals whose true outcome is positive (i.e. \eqn{Y = 1}) accross groups defined by a binary protected attribute.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable
-#' @param group Name of the sensitive attribute
+#' @param group Name of the protected attribute. Must consist of only two groups.
 #' @param probs Predicted probabilities
 #' @param confint Logical indicating whether to calculate confidence intervals
 #' @param bootstraps Number of bootstraps to use for confidence intervals
@@ -46,7 +45,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #'
 #' # Evaluate Balance for Positive Class
 #' eval_pos_class_bal(
@@ -62,10 +61,14 @@
 eval_pos_class_bal <- function(data, outcome, group, probs, confint = TRUE,
                                alpha = 0.05, bootstraps = 2500,
                                digits = 2, message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
+  }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
   }
 
   pos_data <- data[data[[outcome]] == 1, ]

### 2025-08-05 R/eval_pos_pred_parity.R

@@ -1,14 +1,14 @@
 #' Examine Positive Predictive Parity of a Model
 #'
 #' This function evaluates *positive predictive predictive parity*, a key fairness criterion that
-#' compares the *Positive Predictive Value (PPV)* between groups defined by a sensitive attribute.
+#' compares the *Positive Predictive Value (PPV)* between groups defined by a binary protected attribute.
 #' In other words, it assesses whether, among individuals predicted to be positive,
 #' the probability of being truly positive is equal across subgroups.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable, it must be binary
-#' @param group Name of the sensitive attribute
+#' @param group Name of the protected attribute. Must consist of only two groups.
 #' @param probs Name of the predicted outcome variable
 #' @param cutoff Threshold for the predicted outcome, default is 0.5
 #' @param confint Whether to compute 95% confidence interval, default is TRUE
@@ -50,7 +50,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Positive Predictive Parity
@@ -68,11 +68,16 @@
 eval_pos_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                                  bootstraps = 2500, alpha = 0.05,
                                  digits = 2, message = TRUE) {
-  # Check if outcome is binary
-  unique_values <- sort(unique(data[[outcome]]))
+  # Check if outcome and groups are binary
+  unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
   }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
+  }
+
 
   ppv <- get_ppv(
     data = data, outcome = outcome, group = group, probs = probs,

### 2025-08-05 R/eval_pred_equality.R

@@ -1,14 +1,14 @@
 #' Examine Predictive Equality of a Model
 #'
 #' This function evaluates predictive equality, a fairness metric that compares the
-#' False Positive Rate (FPR) between groups defined by a sensitive attribute. It assesses
+#' False Positive Rate (FPR) between groups defined by a binary protected attribute. It assesses
 #' whether individuals from different groups are equally likely to be incorrectly flagged as
 #' positive when they are, in fact, negative.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable, it must be binary
-#' @param group Name of the sensitive attribute
+#' @param group Name of the protected attribute. Must consist of only two groups.
 #' @param probs Name of the predicted outcome variable
 #' @param cutoff Threshold for the predicted outcome, default is 0.5
 #' @param confint Whether to compute 95% confidence interval, default is TRUE
@@ -48,7 +48,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protectedR attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Predictive Equality
@@ -66,11 +66,16 @@
 eval_pred_equality <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                                alpha = 0.05, bootstraps = 2500,
                                digits = 2, message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
   }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
+  }
+
 
   fpr <- get_fpr(
     data = data, outcome = outcome, group = group, probs = probs,

### 2025-08-05 R/eval_stats_parity.R

@@ -1,11 +1,11 @@
 #' Examine Statistical Parity of a Model
 #'
-#' This function assesses *statistical parity* - also known as *demographic parity* - in the predictions of a binary classifier across two groups defined by a sensitive attribute. Statistical parity compares the rate at which different groups receive a positive prediction, irrespective of the true outcome. It reports the Positive Prediction Rate (PPR) for each group, their differences, ratios, and bootstrap-based confidence regions.
+#' This function assesses *statistical parity* - also known as *demographic parity* - in the predictions of a binary classifier across two groups defined by a protected attribute. Statistical parity compares the rate at which different groups receive a positive prediction, irrespective of the true outcome. It reports the Positive Prediction Rate (PPR) for each group, their differences, ratios, and bootstrap-based confidence regions.
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' protected attribute
 #' @param outcome Name of the outcome variable, it must be binary
-#' @param group Name of the sensitive attribute
+#' @param group Name of the protected attribute. Must consist of only two groups.
 #' @param probs Name of the predicted outcome variable
 #' @param cutoff Threshold for the predicted outcome, default is 0.5
 #' @param confint Whether to compute 95% confidence interval, default is TRUE
@@ -47,7 +47,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Statistical Parity
@@ -64,10 +64,14 @@
 eval_stats_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                               bootstraps = 2500, alpha = 0.05, digits = 2,
                               message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
+  }
[…]

### 2025-08-05 R/eval_treatment_equality.R

@@ -1,14 +1,14 @@
 #' Examine Treatment Equality of a Model
 #'
 #' This function evaluates *Treatment Equality*, a fairness criterion that assesses whether the
-#' ratio of false negatives to false positives is similar across groups (e.g., based on gender or race).
+#' ratio of false negatives to false positives is similar across groups defined by a binary protected attribute.
 #' Treatment Equality ensures that the model does not disproportionately favor or disadvantage any group
 #' in terms of the relative frequency of missed detections (false negatives) versus false alarms (false positives).
 #'
 #' @param data Data frame containing the outcome, predicted outcome, and
-#' sensitive attribute
+#' binary protected attribute
 #' @param outcome Name of the outcome variable
-#' @param group Name of the sensitive attribute
+#' @param group group Name of the binary protected attribute. Must consist of only two groups.
 #' @param probs Predicted probabilities
 #' @param cutoff Cutoff value for the predicted probabilities
 #' @param confint Logical indicating whether to calculate confidence intervals
@@ -49,7 +49,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #'
 #' # Evaluate Treatment Equality
 #' eval_treatment_equality(
@@ -71,11 +71,16 @@
 eval_treatment_equality <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                                     alpha = 0.05, bootstraps = 2500,
                                     digits = 2, message = TRUE) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
+  groups <- unique(data[[group]])
   if (!(length(unique_values) == 2 && all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
+    stop("`outcome` must be binary (containing only 0 and 1).")
   }
+  if (!(length(groups) == 2)) {
+    stop("`group` argument must only consist of two groups (i.e. `length(unique(data[[group]])) == 2`")
+  }
+
 
   err_ratio <- get_err_ratio(
     data = data, outcome = outcome, group = group, probs = probs,

### 2025-08-05 R/get_fairness_metrics.R

@@ -1,6 +1,6 @@
 #' Compute Fairness Metrics for Binary Classification
 #'
-#' Computes a comprehensive set of fairness metrics for binary classification models, disaggregated by a sensitive attribute (e.g., race, gender). Optionally, conditional fairness can be evaluated using a second attribute and a specified condition. The function also computes corresponding performance metrics used in the fairness calculations.
+#' Computes a comprehensive set of fairness metrics for binary classification models, disaggregated by a binary protected attribute (e.g., race, gender). Optionally, conditional fairness can be evaluated using a second attribute and a specified condition. The function also computes corresponding performance metrics used in the fairness calculations.
 #'
 #' The results are returned as a list of two data frames:
 #' - `performance`: Contains performance metrics (e.g., TPR, FPR, PPV) by group.
@@ -23,7 +23,7 @@
 #'
 #' @param data A data frame containing the outcome, group, and predicted probabilities.
 #' @param outcome The name of the column containing the true binary outcome.
-#' @param group The name of the column representing the sensitive attribute (e.g., race, gender).
+#' @param group The name of the column representing the binary protected attribute (e.g., race, gender).
 #' @param group2 Define if conditional statistical parity is desired. Name of a secondary group variable used for conditional fairness analysis.
 #' @param condition Define if conditional statistical parity is desired. If the conditional group is categorical, the condition
 #' supplied must be a character of the levels to condition on. If the conditional
@@ -63,7 +63,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Get Fairness Metrics
@@ -93,12 +93,16 @@ get_fairness_metrics <- function(data,
                                   bootstraps = 2500,
                                   alpha = 0.05,
                                   digits = 2) {
-  # Check if outcome is binary
+  # Check if outcome and groups are binary
   unique_values <- unique(data[[outcome]])
-  if (!(length(unique_values) == 2 &&
-        all(unique_values %in% c(0, 1)))) {
-    stop("Outcome must be binary (containing only 0 and 1).")
[…]

### 2025-08-05 utils/MultiGroupFairness.R

@@ -40,7 +40,7 @@
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Maximum-Minimum Difference
@@ -102,7 +102,7 @@ eval_max_min_diff <- function(data, outcome, group, probs, cutoff = 0.5, digits 
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Maximum-Minimum Ratio
@@ -164,7 +164,7 @@ eval_max_min_ratio <- function(data, outcome, group, probs, cutoff = 0.5,
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Max Absolute Difference
@@ -227,7 +227,7 @@ eval_max_abs_diff <- function(data, outcome, group, probs, cutoff = 0.5,
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
-#' # We will use sex as the sensitive attribute and day_28_flg as the outcome.
+#' # We will use sex as the protected attribute and day_28_flg as the outcome.
 #' # We choose threshold = 0.41 so that the overall FPR is around 5%.
 #'
 #' # Evaluate Mean Absolute Deviation
@@ -291,7 +291,7 @@ eval_mean_abs_dev <- function(data, outcome, group, probs, cutoff = 0.5,
 #' test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 #'
 #' # Fairness evaluation
[…]

### 2025-08-06 R/ModelPerformance.R
rewriting vingette

@@ -18,8 +18,8 @@ get_tpr <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
   groups <- sort(unique(data[, group])) # Get sorted unique groups
 
   for (i in groups) {
-    tp <- sum(data[, outcome] == 1 & data[, group] == i & data[, probs] >= cutoff)
-    p <- sum(data[, outcome] == 1 & data[, group] == i)
+    tp <- sum(data[, outcome] == 1 & data[, group] == i & data[, probs] >= cutoff, na.rm=TRUE)
+    p <- sum(data[, outcome] == 1 & data[, group] == i, na.rm = TRUE)
     tpr <- c(tpr, round(tp / p, digits)) # Calculate TPR and add to vector
   }
 
@@ -44,8 +44,8 @@ get_fpr <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
   for (i in groups) {
     fp <- sum(data[, outcome] == 0 &
       data[, group] == i &
-      data[, probs] >= cutoff)
-    n <- sum(data[, outcome] == 0 & data[, group] == i)
+      data[, probs] >= cutoff, na.rm = TRUE)
+    n <- sum(data[, outcome] == 0 & data[, group] == i, na.rm=TRUE)
     fpr <- c(fpr, round(fp / n, digits))
   }
   return(fpr)
@@ -68,9 +68,10 @@ get_ppr <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
   for (i in groups) {
     pp <- sum(
       data[, group] == i &
-        data[, probs] >= cutoff
+        data[, probs] >= cutoff,
+      na.rm = TRUE
     )
-    n <- sum(data[, group] == i)
+    n <- sum(data[, group] == i, na.rm=TRUE)
     ppr <- c(ppr, round(pp / n, digits))
   }
   return(ppr)
@@ -93,9 +94,9 @@ get_ppv <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
   for (i in groups) {
     tp <- sum(data[, outcome] == 1 &
       data[, group] == i &
-      data[, probs] >= cutoff)
+      data[, probs] >= cutoff, na.rm=TRUE)
     pp <- sum(data[, group] == i &
-      data[, probs] >= cutoff)
+      data[, probs] >= cutoff, na.rm=TRUE)
     ppv <- c(ppv, round(tp / pp, digits))
   }
   return(ppv)
[…]

### 2025-08-06 R/eval_stats_parity.R

@@ -75,9 +75,14 @@ eval_stats_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint
   }
 
   ppr <- get_ppr(
-    data = data, outcome = outcome, group = group, probs = probs,
-    cutoff = cutoff, digits = digits
+    data = data,
+    outcome = outcome,
+    group = group,
+    probs = probs,
+    cutoff = cutoff,
+    digits = digits
   )
+
 
   ppr_diff <- ppr[[1]] - ppr[[2]]
   ppr_ratio <- ppr[[1]] / ppr[[2]]

### 2025-08-06 R/ModelPerformance.R
rewriting vingette

@@ -142,13 +142,10 @@ get_acc <- function(data, outcome, group, probs, cutoff = 0.5, digits = 2) {
   acc <- c()
   groups <- sort(unique(data[, group]))
   for (i in groups) {
-    tp <- sum(data[, outcome] == 1 &
-      data[, group] == i &
-      data[, probs] >= cutoff, na.rm=TRUE)
-    tn <- sum(data[, outcome] == 0 &
-      data[, group] == i &
-      data[, probs] < cutoff)
-    p <- sum(data[, group] == i, na.rm = TRUE)
+    groups <- sort(unique(data[[group]]))
+    tp <- sum(data[, outcome] == 1 & data[,group] == i & data[,probs] >= cutoff, na.rm=TRUE)
+    tn <- sum(data[, outcome] == 0 & data[,group] == i & data[,probs] < cutoff, na.rm=TRUE)
+    p <- sum(data[,group] == i, na.rm = TRUE)
     acc <- c(acc, round((tp + tn) / p, digits))
   }
   return(acc)
@@ -215,7 +212,7 @@ get_avg_prob <- function(data, outcome, group, probs, digits = 2) {
   groups <- sort(unique(data[, group]))
   for (i in groups) {
     sub_data <- data[data[, group] == i, ]
-    avg_prob <- c(avg_prob, round(mean(sub_data[, probs]), digits))
+    avg_prob <- c(avg_prob, round(mean(sub_data[, probs],na.rm=TRUE), digits))
   }
   return(avg_prob)
 }

### 2025-08-06 R/get_fairness_metrics.R

@@ -398,7 +398,7 @@ get_fairness_metrics <- function(data,
         digits = digits
       )
 
-      if (!(is.null(group2) & is.null(condition))) {
+      if (!is.null(group2) && !is.null(condition)){
         # Getting Diff and Ratio Vals
         diff_vals <- list(
           ppr_diff = ppr[[1]] - ppr[[2]],

### 2025-08-07 utils/joss_paper/paper.Rmd
dynamically renaming CI columns

@@
 $performance
-                                    Metric Group1 Group2
+                                    Metric GroupFemale GroupMale
 1                 Positive Prediction Rate        0.17      0.08
@@
 $fairness
-                           Metric Difference        Diff_CI Ratio     Ratio_CI
+                           Metric Difference    95% Diff CI Ratio 95% Ratio CI
 1              Statistical Parity       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]

### 2025-08-07 R/eval_acc_parity.R

@@ -118,9 +118,9 @@ eval_acc_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint =
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),
       "Difference",
-      "95% Diff CI",
+      paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
-      "95% Ratio CI"
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
 

### 2025-08-07 R/eval_bs_parity.R

@@ -114,9 +114,9 @@ eval_bs_parity <- function(data, outcome, group, probs, confint = TRUE,
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),
       "Difference",
-      "95% Diff CI",
+      paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
-      "95% Ratio CI"
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
 

### 2025-08-07 R/eval_eq_odds.R

@@ -166,7 +166,10 @@ eval_eq_odds <- function(data, outcome, group, probs, cutoff = 0.5, confint = TR
       "Metric",
       paste0("Group ", sort(unique(data[[group]]))[[1]]),
       paste0("Group ", sort(unique(data[[group]]))[[2]]),
-      "Difference", "95% Diff CI", "Ratio", "95% Ratio CI"
+      "Difference", 
+      paste0((1-alpha)*100, "% Diff CI"),
+      "Ratio",
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
     # Print summary message if desired

### 2025-08-07 R/eval_eq_opp.R

@@ -33,9 +33,9 @@
 #'   \item \code{Group1} Estimated FNR and FPR for the first group.
 #'   \item \code{Group2} Estimated FNR and FPR for the second group.
 #'   \item \code{Difference} The difference in FNR between the two groups, computed as the FNR of Group1 minus the FNR of Group2.
-#'   \item \code{95\% Diff CI} The (1 - \code{alpha}) confidence interval for the FNR difference.
+#'   \item \code{1-alpha\% Diff CI} The (1 - \code{alpha}) confidence interval for the FNR difference.
 #'   \item \code{Ratio} The ratio of FNRs between Group1 and Group2, computed as FNR for Group1 divided by FNR for Group2.
-#'   \item \code{95\% Ratio CI} The corresponding confidence interval for the FNR ratio.
+#'   \item \code{1-alpha\% Ratio CI} The corresponding confidence interval for the FNR ratio.
 #' }
 #'
 #' @importFrom stats qnorm sd
@@ -139,7 +139,10 @@ eval_eq_opp <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRU
     colnames(results_df) <- c(
       "Metric", paste0("Group", sort(unique(data[[group]]))[[1]]),
       paste0("Group", sort(unique(data[[group]]))[[2]]),
-      "Difference", "95% Diff CI", "Ratio", "95% Ratio CI"
+      "Difference", 
+      paste0((1-alpha)*100, "% Diff CI"),
+      "Ratio",
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
     # Print message if desired

### 2025-08-07 R/eval_neg_class_bal.R

@@ -118,9 +118,9 @@ eval_neg_class_bal <- function(data, outcome, group, probs, confint = TRUE,
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),
       "Difference",
-      "95% Diff CI",
+      paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
-      "95% Ratio CI"
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
     if (message) {

### 2025-08-07 R/eval_neg_pred_parity.R

@@ -120,9 +120,9 @@ eval_neg_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, conf
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),
       "Difference",
-      "95% Diff CI",
+      paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
-      "95% Ratio CI"
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
 

### 2025-08-07 R/eval_pos_pred_parity.R

@@ -121,9 +121,9 @@ eval_pos_pred_parity <- function(data, outcome, group, probs, cutoff = 0.5, conf
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),
       "Difference",
-      "95% Diff CI",
+      paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
-      "95% Ratio CI"
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
     if (message) {

### 2025-08-07 R/eval_pred_equality.R

@@ -120,9 +120,9 @@ eval_pred_equality <- function(data, outcome, group, probs, cutoff = 0.5, confin
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),
       "Difference",
-      "95% Diff CI",
+      paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
-      "95% Ratio CI"
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
 

### 2025-08-07 R/eval_stats_parity.R

@@ -124,7 +124,10 @@ eval_stats_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint
       "Metric",
       paste0("Group", sort(unique(data[[group]]))[[1]]),
       paste0("Group", sort(unique(data[[group]]))[[2]]),
-      "Difference", "95% Diff CI", "Ratio", "95% Ratio CI"
+      "Difference", 
+      paste0((1-alpha)*100, "% Diff CI"),
+      "Ratio",
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
 

### 2025-08-07 R/eval_treatment_equality.R

@@ -129,9 +129,9 @@ eval_treatment_equality <- function(data, outcome, group, probs, cutoff = 0.5, c
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),
       "Difference",
-      "95% Diff CI",
+      paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
-      "95% Ratio CI"
+      paste0((1-alpha)*100, "% Ratio CI")
     )
 
     if (message) {

### 2025-08-07 R/get_fairness_metrics.R

@@ -537,7 +537,18 @@ get_fairness_metrics <- function(data,
         row.names = NULL
       )
 
-
+      colnames(model_summary) <- c(
+        "Metric", 
+        paste0("Group", sort(unique(data[[group]]))[1]), 
+        paste0("Group", sort(unique(data[[group]]))[2])
+        )
+      colnames(fairness_summary) <- c(
+        "Metric",
+        "Difference",
+        paste0((1 - alpha) * 100, "% Diff CI"),
+        "Ratio",
+        paste0((1 - alpha) * 100, "% Ratio CI")
+      )
     } else{
       model_summary <- data.frame(
         Metric = c(
@@ -581,6 +592,12 @@ get_fairness_metrics <- function(data,
         row.names = NULL
 
       )
+
+      colnames(model_summary) <- c(
+        "Metric", 
+        paste0("Group", sort(unique(data[[group]]))[1]), 
+        paste0("Group", sort(unique(data[[group]]))[2])
+        )
 
     }
   } else{

### 2025-08-11 R/eval_cond_stats_parity.R
removing conditional statistical parity from the package and from get_fairness_metrics

@@ -1,129 +1,0 @@
-#' Examine Conditional Statistical Parity of a Model
-#'
-#' This function evaluates *conditional statistical parity*, which measures fairness by comparing positive prediction rates across two groups defined by a binary protected attribute within a defined subgroup of the population. This is useful in scenarios where fairness should be evaluated in a more context-specific way—e.g., within a particular hospital unit or age bracket. Conditional statistical parity is a refinement of standard statistical parity. Instead of comparing prediction rates across groups in the entire dataset, it restricts the comparison
-#' to a specified subset of the population, defined by a conditioning variable.
-#'
-#' The function supports both categorical and continuous conditioning variables. For continuous variables, you can supply a threshold expression like `"<50"` or `">=75"` to the \code{condition} parameter.
-#'
-#' @param data Data frame containing the outcome, predicted outcome, and
-#' binary protected attribute
-#' @param outcome Name of the outcome variable, it must be binary
-#' @param group Name of the binary protected attribute. Must consist of only two groups.
-#' @param group2 Name of the group to condition on
-#' @param condition If the conditional group is categorical, the condition
-#' supplied must be a character of the levels to condition on. If the conditional
-#' group is continuous, the conditions supplied must be a character containing
-#' the sign of the condition and the value to threshold the continuous variable
-#' (e.g. "<50", ">50", "<=50", ">=50").
-#' @param probs Name of the predicted outcome variable
-#' @param confint Whether to compute 95% confidence interval, default is TRUE
-#' @param cutoff Threshold for the predicted outcome, default is 0.5
-#' @param bootstraps Number of bootstrap samples, default is 2500
-#' @param alpha The 1 - significance level for the confidence interval, default is 0.05
-#' @param digits Number of digits to round the results to, default is 2
-#' @param message Logical; if TRUE (default), prints a textual summary of the
-#' fairness evaluation. Only works if `confint` is TRUE.
-#' @return A list containing the following elements:
-#'  - Conditions: The conditions used to calculate the conditional PPR
-#'  - PPR_Group1: Positive Prediction Rate for the first group
[…]

### 2025-08-11 R/eval_stats_parity.R

@@ -59,7 +59,6 @@
 #'   cutoff = 0.41
 #' )
 #' }
-#' @seealso \code{\link{eval_cond_stats_parity}}
 #' @export
 eval_stats_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint = TRUE,
                               bootstraps = 2500, alpha = 0.05, digits = 2,
@@ -124,7 +123,7 @@ eval_stats_parity <- function(data, outcome, group, probs, cutoff = 0.5, confint
       "Metric",
       paste0("Group", sort(unique(data[[group]]))[[1]]),
       paste0("Group", sort(unique(data[[group]]))[[2]]),
-      "Difference", 
+      "Difference",
       paste0((1-alpha)*100, "% Diff CI"),
       "Ratio",
       paste0((1-alpha)*100, "% Ratio CI")

### 2025-08-11 R/get_fairness_metrics.R

@@ -1,6 +1,6 @@
 #' Compute Fairness Metrics for Binary Classification
 #'
-#' Computes a comprehensive set of fairness metrics for binary classification models, disaggregated by a binary protected attribute (e.g., race, gender). Optionally, conditional fairness can be evaluated using a second attribute and a specified condition. The function also computes corresponding performance metrics used in the fairness calculations.
+#' Computes a comprehensive set of fairness metrics for binary classification models, disaggregated by a binary protected attribute. The function also computes corresponding performance metrics used in the fairness calculations.
 #'
 #' The results are returned as a list of two data frames:
 #' - `performance`: Contains performance metrics (e.g., TPR, FPR, PPV) by group.
@@ -8,8 +8,6 @@
 #'
 #' ### Fairness Metrics Included:
 #' - **Statistical Parity**: Difference in positive prediction rates across groups.
-#' - **Conditional Statistical Parity** *(if group2 and condition are specified)*:
-#'   Parity conditioned on a second group and value.
 #' - **Equal Opportunity**: Difference in true positive rates (TPR) across groups.
 #' - **Predictive Equality**: Difference in false positive rates (FPR) across groups.
 #' - **Balance for Positive Class**: Checks whether the predicted probability distributions for
@@ -24,12 +22,6 @@
 #' @param data A data frame containing the outcome, group, and predicted probabilities.
 #' @param outcome The name of the column containing the true binary outcome.
 #' @param group The name of the column representing the binary protected attribute (e.g., race, gender).
-#' @param group2 Define if conditional statistical parity is desired. Name of a secondary group variable used for conditional fairness analysis.
-#' @param condition Define if conditional statistical parity is desired. If the conditional group is categorical, the condition
-#' supplied must be a character of the levels to condition on. If the conditional
-#' group is continuous, the conditions supplied must be a character containing
[…]

### 2025-08-11 utils/joss_paper/paper.Rmd
updating vingette and  JOSS article

--- 2025-08-04 utils/joss_paper/paper.Rmd
@@
 
-- **Independence:** Statistical Parity (compares the overall rate of positive predictions between groups), Conditional Statistical Parity (compares the rate of positive predictions between groups within a specific subgroup).
+- **Independence:** Statistical Parity (compares the overall rate of positive predictions between groups).
 
@@
 In this example, we evaluate the model's fairness with respect to the binary protected attribute `gender`.
-For conditional statistical parity, we condition on age greater than 60 years.
 The model is trained on a subset of the data and the predictions are made and evaluated on a test set.
@@
  group = "gender",
- group2 = "age",
- condition = ">=60",
  probs = "pred",
@@
 $performance
-                                    Metric Group1 Group2
+                                    Metric GroupFemale GroupMale
 1                 Positive Prediction Rate        0.17      0.08
-2     Conditional Positive Prediction Rate   0.34   0.21
-3                      False Negative Rate   0.38   0.62
-4                      False Positive Rate   0.08   0.03
-5            Avg. Predicted Positive Prob.   0.46   0.37
-6            Avg. Predicted Negative Prob.   0.15   0.10
-7                Positive Predictive Value   0.62   0.66
-8                Negative Predictive Value   0.92   0.90
-9                              Brier Score   0.09   0.08
-10                                Accuracy   0.87   0.88
-11 (False Negative)/(False Positive) Ratio   1.03   3.24
+2                      False Negative Rate        0.38      0.62
+3                      False Positive Rate        0.08      0.03
+4            Avg. Predicted Positive Prob.        0.46      0.37
+5            Avg. Predicted Negative Prob.        0.15      0.10
+6                Positive Predictive Value        0.62      0.66
+7                Negative Predictive Value        0.92      0.90
+8                              Brier Score        0.09      0.08
+9                                Accuracy        0.87      0.88
+10 (False Negative)/(False Positive) Ratio        1.03      3.24
 
 $fairness
-                           Metric Difference        Diff_CI Ratio     Ratio_CI
+                           Metric Difference    95% Diff CI Ratio 95% Ratio CI
 1              Statistical Parity       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
-2  Conditional Statistical Parity       0.13   [0.05, 0.21]  1.62 [1.17, 2.24]
-3               Equal Opportunity      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
-4             Predictive Equality       0.05   [0.02, 0.08]  2.67 [1.38, 5.15]
-5      Balance for Positive Class       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
-6      Balance for Negative Class       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
-7      Positive Predictive Parity      -0.04  [-0.21, 0.13]  0.94 [0.71, 1.24]
-8      Negative Predictive Parity       0.02  [-0.02, 0.06]  1.02 [0.98, 1.07]
-9              Brier Score Parity       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
-10        Overall Accuracy Parity      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
-11             Treatment Equality      -2.21  [-4.44, 0.02]  0.32  [0.14, 0.7]
+2               Equal Opportunity      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
+3             Predictive Equality       0.05   [0.02, 0.08]  2.67 [1.38, 5.15]
+4      Balance for Positive Class       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
+5      Balance for Negative Class       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
+6      Positive Predictive Parity      -0.04  [-0.21, 0.13]  0.94 [0.71, 1.24]
+7      Negative Predictive Parity       0.02  [-0.02, 0.06]  1.02 [0.98, 1.07]
+8              Brier Score Parity       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
+9         Overall Accuracy Parity      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
+10             Treatment Equality      -2.21  [-4.44, 0.02]  0.32  [0.14, 0.7]
 ```

### 2025-08-11 tests/testthat/helper-match-metrics.R
updating tests

@@ -19,8 +19,6 @@ get_fairness_metrics(
   data = test_data,
   outcome = "day_28_flg",
   group = "gender",
-  group2 = "service_unit",
-  condition = "MICU",
   probs = "pred",
   cutoff = 0.41,
   confint = FALSE

### 2025-08-11 tests/testthat/test-Fairness-match-metrics.R

@@ -37,27 +37,6 @@ test_that('eval_bs_parity match metrics test', {
   })
 })
 
-test_that('eval_cond_stats_parity metrics test', {
-  expect_equal({
-    source("helper-match-metrics.R")
-    res <- eval_cond_stats_parity(
-      data = test_data,
-      outcome = "day_28_flg",
-      group = "gender",
-      group2 = "service_unit",
-      condition = "MICU",
-      probs = "pred",
-      confint = FALSE,
-      message = FALSE
-    )[1:3]
-    rownames(res)<- 1
-  },
-  {
-    res <- subset(performance, Metric =="Positive Prediction Rate")[2,]
-    rownames(res)<- 1
-  })
-})
-
 
 test_that('eval_eq_opp match metrics test', {
   expect_equal({

### 2025-08-11 tests/testthat/test-Fairness-no-errors.R

@@ -33,24 +33,6 @@ test_that('eval_bs_parity test', {
   })
 })
 
-test_that('eval_cond_stats_parity test', {
-  expect_no_error({
-  source("helper.R")
-  eval_cond_stats_parity(
-    data = test_data,
-    outcome = "day_28_flg",
-    group = "gender",
-    group2 = "service_unit",
-    condition = "MICU",
-    probs = "pred",
-    cutoff = 0.41,
-    bootstraps = 2500,
-    alpha = 0.05,
-    digits = 2,
-    message = FALSE
-  )
-})
-})
 
 test_that('eval_eq_odds test',{
   expect_no_error({

### 2025-08-11 tests/testthat/test-get_fairness_metrics-no-errors.R

@@ -5,8 +5,6 @@ test_that('get_fairness_metrics CI test', {
       data = test_data,
       outcome = "day_28_flg",
       group = "gender",
-      group2 = "age",
-      condition = ">=60",
       probs = "pred",
       confint = TRUE,
       cutoff = 0.41,
@@ -25,8 +23,6 @@ test_that('get_fairness_metrics no CI test', {
       data = test_data,
       outcome = "day_28_flg",
       group = "gender",
-      group2 = "age",
-      condition = ">=60",
       probs = "pred",
       confint = FALSE,
       cutoff = 0.41,

### 2025-08-21 utils/joss_paper/paper.Rmd
updating get_fairness_metrics to return a dataframe

@@
 output:
-  # rticles::joss_article:
-  #   keep_tex: true
+  rticles::joss_article:
+    keep_tex: true
   md_document:
@@
 
-$performance
-                                    Metric GroupFemale GroupMale
-1                 Positive Prediction Rate        0.17      0.08
-2                      False Negative Rate        0.38      0.62
-3                      False Positive Rate        0.08      0.03
-4            Avg. Predicted Positive Prob.        0.46      0.37
-5            Avg. Predicted Negative Prob.        0.15      0.10
-6                Positive Predictive Value        0.62      0.66
-7                Negative Predictive Value        0.92      0.90
-8                              Brier Score        0.09      0.08
-9                                Accuracy        0.87      0.88
-10 (False Negative)/(False Positive) Ratio        1.03      3.24
+           Fairness Assesment                                  Metric
+1          Statistical Parity                Positive Prediction Rate
+2           Equal Opportunity                     False Negative Rate
+3         Predictive Equality                     False Positive Rate
+4  Balance for Positive Class           Avg. Predicted Positive Prob.
+5  Balance for Negative Class           Avg. Predicted Negative Prob.
+6  Positive Predictive Parity               Positive Predictive Value
+7  Negative Predictive Parity               Negative Predictive Value
+8          Brier Score Parity                             Brier Score
+9     Overall Accuracy Parity                                Accuracy
+10         Treatment Equality (False Negative)/(False Positive) Ratio
 
-$fairness
-                           Metric Difference    95% Diff CI Ratio 95% Ratio CI
-1              Statistical Parity       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
-2               Equal Opportunity      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
-3             Predictive Equality       0.05   [0.02, 0.08]  2.67 [1.38, 5.15]
-4      Balance for Positive Class       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
-5      Balance for Negative Class       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
-6      Positive Predictive Parity      -0.04  [-0.21, 0.13]  0.94 [0.71, 1.24]
-7      Negative Predictive Parity       0.02  [-0.02, 0.06]  1.02 [0.98, 1.07]
-8              Brier Score Parity       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
-9         Overall Accuracy Parity      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
-10             Treatment Equality      -2.21  [-4.44, 0.02]  0.32  [0.14, 0.7]
+   GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
+1         0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.49, 3.04]
+2         0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
+3         0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.4, 5.08]
+4         0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
+5         0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
+6         0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.22]
+7         0.92      0.90       0.02  [-0.02, 0.06]  1.02 [0.98, 1.07]
+8         0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
+9         0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
+10        1.03      3.24      -2.21 [-4.38, -0.04]  0.32 [0.15, 0.68]
 ```

### 2025-08-21 utils/joss_paper/paper.tex
updating get_fairness_metrics to return a dataframe

\documentclass[10pt,a4paper,onecolumn]{article}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{tikz}
\usepackage{hyperref}
\hypersetup{colorlinks,breaklinks,
            urlcolor=[rgb]{0.0, 0.5, 1.0},
            linkcolor=[rgb]{0.0, 0.5, 1.0}}
\usepackage{caption}
\usepackage{tcolorbox}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{seqsplit}
\usepackage{fixltx2e} % provides \textsubscript
\usepackage[
  backend=biber,
%  style=alphabetic,
%  citestyle=numeric
]{biblatex}
\bibliography{paper.bib}

% --- Page layout -------------------------------------------------------------
\usepackage[top=3.5cm, bottom=3cm, right=1.5cm, left=1.0cm,
            headheight=2.2cm, reversemp, includemp, marginparwidth=4.5cm]{geometry}

% --- Default font ------------------------------------------------------------
% \renewcommand\familydefault{\sfdefault}

% --- Style -------------------------------------------------------------------
\renewcommand{\bibfont}{\small \sffamily}
\renewcommand{\captionfont}{\small\sffamily}
\renewcommand{\captionlabelfont}{\bfseries}

% --- Section/SubSection/SubSubSection ----------------------------------------
\titleformat{\section}
  {\normalfont\sffamily\Large\bfseries}
  {}{0pt}{}
\titleformat{\subsection}
  {\normalfont\sffamily\large\bfseries}
  {}{0pt}{}
\titleformat{\subsubsection}
  {\normalfont\sffamily\bfseries}
  {}{0pt}{}
\titleformat*{\paragraph}
  {\sffamily\normalsize}

% --- Header / Footer ---------------------------------------------------------
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\renewcommand{\headrulewidth}{0.50pt}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{C:/Users/ben29/AppData/Local/R/win-library/4.4/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
\fancyhead[C]{}
\fancyhead[R]{}
\renewcommand{\footrulewidth}{0.25pt}

\fancyfoot[L]{\footnotesize{\sffamily , (). fairmetrics: An R package
for group fairness
evaluation. \textit{Journal of Open Source Software}, (), . \href{https://doi.org/}{https://doi.org/}}}

\fancyfoot[R]{\sffamily \thepage}
\makeatletter
\let\ps@plain\ps@fancy
\fancyheadoffset[L]{4.5cm}
\fancyfootoffset[L]{4.5cm}

% --- Macros ---------

\definecolor{linky}{rgb}{0.0, 0.5, 1.0}

\newtcolorbox{repobox}
   {colback=red, colframe=red!75!black,
     boxrule=0.5pt, arc=2pt, left=6pt, right=6pt, top=3pt, bottom=3pt}

\newcommand{\ExternalLink}{%
   \tikz[x=1.2ex, y=1.2ex, baseline=-0.05ex]{%
       \begin{scope}[x=1ex, y=1ex]
           \clip (-0.1,-0.1)
               --++ (-0, 1.2)
               --++ (0.6, 0)
               --++ (0, -0.6)
               --++ (0.6, 0)
               --++ (0, -1);
           \path[draw,
               line width = 0.5,
               rounded corners=0.5]
               (0,0) rectangle (1,1);
       \end{scope}
       \path[draw, line width = 0.5] (0.5, 0.5)
           -- (1, 1);
       \path[draw, line width = 0.5] (0.6, 1)
           -- (1, 1) -- (1, 0.6);
       }
   }

% --- Title / Authors ---------------------------------------------------------
% patch \maketitle so that it doesn't center
\patchcmd{\@maketitle}{center}{flushleft}{}{}
\patchcmd{\@maketitle}{center}{flushleft}{}{}
% patch \maketitle so that the font size for the title is normal
\patchcmd{\@maketitle}{\LARGE}{\LARGE\sffamily}{}{}
% patch the patch by authblk so that the author block is flush left
\def\maketitle{{%
  \renewenvironment{tabular}[2][]
    {\begin{flushleft}}
    {\end{flushleft}}
  \AB@maketitle}}
\makeatletter
\renewcommand\AB@affilsepx{ \protect\Affilfont}
%\renewcommand\AB@affilnote[1]{{\bfseries #1}\hspace{2pt}}
\renewcommand\AB@affilnote[1]{{\bfseries #1}\hspace{3pt}}
\makeatother
\renewcommand\Authfont{\sffamily\bfseries}
\renewcommand\Affilfont{\sffamily\small\mdseries}
\setlength{\affilsep}{1em}

\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}

\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}

\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}

\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={fairmetrics: An R package for group fairness evaluation},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
\makeatother
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
%From Pandoc 3.1.8
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{fairmetrics: An R package for group fairness evaluation}

        \author[1]{Benjamin Smith}
          \author[1]{Jianhui Gao}
          \author[1]{Benson Chou}
          \author[1]{Jessica Gronsbell}

      \affil[1]{Department of Statistical Science, University of
Toronto}
  \date{\vspace{-5ex}}

\begin{document}
\maketitle

\marginpar{
  %\hrule
  \sffamily\small

  {\bfseries DOI:} \href{https://doi.org/}{\color{linky}{}}

  \vspace{2mm}

  {\bfseries Software}
  \begin{itemize}
    \setlength\itemsep{0em}
    \item \href{}{\color{linky}{Review}} \ExternalLink
    \item \href{}{\color{linky}{Repository}} \ExternalLink
    \item \href{}{\color{linky}{Archive}} \ExternalLink
  \end{itemize}

  \vspace{2mm}

  {\bfseries Submitted:} \\
  {\bfseries Published:}

  \vspace{2mm}
  {\bfseries License}\\
  Authors of papers retain copyright and release the work under a Creative Commons Attribution 4.0 International License (\href{http://creativecommons.org/licenses/by/4.0/}{\color{linky}{CC-BY}}).
}

\section{Summary}\label{summary}

Fairness is a growing area of machine learning (ML) that focuses on
ensuring that models do not produce systematically biased outcomes
across groups defined by protected attributes, such as race, gender, or
age. The \texttt{fairmetrics} R package provides a user-friendly
framework for rigorously evaluating group-based fairness criteria,
including independence (e.g., statistical parity), separation (e.g.,
equalized odds), and sufficiency (e.g., predictive parity) for binary
protected attributes. The package provides both point and interval
estimates for a variety of commonly used criteria. \texttt{fairmetrics}
also includes an example dataset derived from the Medical Information
Mart for Intensive Care, version II (MIMIC-II) database (Goldberger et
al. 2000; J. Raffa 2016) to demonstrate its use.

\section{Statement of Need}\label{statement-of-need}

ML models are increasingly used in high-stakes domains such as criminal
justice, healthcare, finance, employment, and education (Mehrabi et al.
2021; Mattu 2016; Gao et al. 2024). Existing fairness evaluation
software report point estimates and/or visualizations, without any
measures of uncertainty. This limits users' ability to determine whether
observed disparities are statistically significant. \texttt{fairmetrics}
addresses this limitation by including confidence intervals for both
difference and ratio based fairness metrics to enable more robust and
statistically grounded fairness assessments.

\section{Fairness Criteria}\label{fairness-criteria}

\texttt{fairmetrics} is designed to evaluate fairness of binary
classification models across binary protected attributes. The package
supports the evaluation of metrics belonging to three major group
fairness criteria:

\begin{itemize}
\item
  \textbf{Independence:} Statistical Parity (compares the overall rate
  of positive predictions between groups).
\item
  \textbf{Separation:} Equal Opportunity (compares false negative rates
  between groups), Predictive Equality (compares false positive rates
  between groups), Balance for Positive Class (compares the average
  predicted probabilities among individuals whose true outcome is
  positive across groups), and Balance for Negative Class (compares the
  average predicted probabilities among individuals whose true outcome
  is negative across groups).
\item
  \textbf{Sufficiency:} Positive Predictive Parity (compares the
  positive predictive values across groups), Negative Predictive Parity
  (compares the negative predictive values across groups).
\end{itemize}

The package also includes additional metrics, such as the Brier Score
Parity (compares the Brier score across groups), Accuracy Parity
(compares the overall accuracy across groups), and Treatment Equality
(compares the ratio of false negatives to false positives across
groups).

\section{Evaluating Fairness
Criteria}\label{evaluating-fairness-criteria}

The input required to evaluate model fairness with the
\texttt{fairmetrics} package is a \texttt{data.frame} or \texttt{tibble}
containing the model's predicted probabilities, the true outcomes, and
the protected attribute. \hyperref[workflow]{Figure ~\ref*{workflow}}
shows the workflow for using \texttt{fairmetrics}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{fairmetrics-workflow.png}}
\caption{Workflow for using \texttt{fairmetrics} to evaluate model
fairness across multiple criteria. \label{workflow}}
\end{figure}

A simple example of how to use the \texttt{fairmetrics} package is
illustrated below. The example makes use of the
\texttt{mimic\_preprocessed} dataset, a pre-processed version of the
Indwelling Arterial Catheter (IAC) Clinical dataset, from the MIMIC-II
clinical database (Goldberger et al. 2000; J. Raffa 2016; J. D. Raffa et
al. 2016).

While the choice of fairness metric used is context dependent, we show
all criteria available with the \texttt{get\_fairness\_metrics()}
function for illustrative purposes. In this example, we evaluate the
model's fairness with respect to the binary protected attribute
\texttt{gender}. The model is trained on a subset of the data and the
predictions are made and evaluated on a test set. A statistically
significant difference across groups at a given level of significance is
indicated when the confidence interval for a difference-based metric
does not include zero or when the interval for a ratio-based metric does
not include one.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Train a classification model (e.g., random forest).}
\CommentTok{\# Add the vector of predicted probabilities to the test data}
\CommentTok{\# to evaluate fairness.}
\FunctionTok{library}\NormalTok{(fairmetrics)}
\CommentTok{\# Setting alpha=0.05 for 95\% confidence intervals}
\FunctionTok{get\_fairness\_metrics}\NormalTok{(}
 \AttributeTok{data =}\NormalTok{ test\_data,}
 \AttributeTok{outcome =} \StringTok{"day\_28\_flg"}\NormalTok{,}
 \AttributeTok{group =} \StringTok{"gender"}\NormalTok{,}
 \AttributeTok{probs =} \StringTok{"pred"}\NormalTok{,}
 \AttributeTok{cutoff =} \FloatTok{0.41}\NormalTok{, }
 \AttributeTok{alpha =} \FloatTok{0.05}
\NormalTok{)}

\NormalTok{           Fairness Assesment                                  Metric}
\DecValTok{1}\NormalTok{          Statistical Parity                Positive Prediction Rate}
\DecValTok{2}\NormalTok{           Equal Opportunity                     False Negative Rate}
\DecValTok{3}\NormalTok{         Predictive Equality                     False Positive Rate}
\DecValTok{4}\NormalTok{  Balance }\ControlFlowTok{for}\NormalTok{ Positive Class           Avg. Predicted Positive Prob.}
\DecValTok{5}\NormalTok{  Balance }\ControlFlowTok{for}\NormalTok{ Negative Class           Avg. Predicted Negative Prob.}
\DecValTok{6}\NormalTok{  Positive Predictive Parity               Positive Predictive Value}
\DecValTok{7}\NormalTok{  Negative Predictive Parity               Negative Predictive Value}
\DecValTok{8}\NormalTok{          Brier Score Parity                             Brier Score}
\DecValTok{9}\NormalTok{     Overall Accuracy Parity                                Accuracy}
\DecValTok{10}\NormalTok{         Treatment }\FunctionTok{Equality}\NormalTok{ (False Negative)}\SpecialCharTok{/}\NormalTok{(False Positive) Ratio}

\NormalTok{   GroupFemale GroupMale Difference    }\DecValTok{95}\SpecialCharTok{\% Diff CI Ratio 95\%}\NormalTok{ Ratio CI}
\DecValTok{1}         \FloatTok{0.17}      \FloatTok{0.08}       \FloatTok{0.09}\NormalTok{   [}\FloatTok{0.05}\NormalTok{, }\FloatTok{0.13}\NormalTok{]  }\FloatTok{2.12}\NormalTok{ [}\FloatTok{1.49}\NormalTok{, }\FloatTok{3.04}\NormalTok{]}
\DecValTok{2}         \FloatTok{0.38}      \FloatTok{0.62}      \SpecialCharTok{{-}}\FloatTok{0.24}\NormalTok{ [}\SpecialCharTok{{-}}\FloatTok{0.39}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.09}\NormalTok{]  }\FloatTok{0.61}\NormalTok{ [}\FloatTok{0.44}\NormalTok{, }\FloatTok{0.86}\NormalTok{]}
\DecValTok{3}         \FloatTok{0.08}      \FloatTok{0.03}       \FloatTok{0.05}\NormalTok{   [}\FloatTok{0.02}\NormalTok{, }\FloatTok{0.08}\NormalTok{]  }\FloatTok{2.67}\NormalTok{  [}\FloatTok{1.4}\NormalTok{, }\FloatTok{5.08}\NormalTok{]}
\DecValTok{4}         \FloatTok{0.46}      \FloatTok{0.37}       \FloatTok{0.09}\NormalTok{   [}\FloatTok{0.04}\NormalTok{, }\FloatTok{0.14}\NormalTok{]  }\FloatTok{1.24}\NormalTok{ [}\FloatTok{1.09}\NormalTok{, }\FloatTok{1.42}\NormalTok{]}
\DecValTok{5}         \FloatTok{0.15}      \FloatTok{0.10}       \FloatTok{0.05}\NormalTok{   [}\FloatTok{0.03}\NormalTok{, }\FloatTok{0.07}\NormalTok{]  }\FloatTok{1.50}\NormalTok{ [}\FloatTok{1.29}\NormalTok{, }\FloatTok{1.74}\NormalTok{]}
\DecValTok{6}         \FloatTok{0.62}      \FloatTok{0.66}      \SpecialCharTok{{-}}\FloatTok{0.04}\NormalTok{  [}\SpecialCharTok{{-}}\FloatTok{0.21}\NormalTok{, }\FloatTok{0.13}\NormalTok{]  }\FloatTok{0.94}\NormalTok{ [}\FloatTok{0.72}\NormalTok{, }\FloatTok{1.22}\NormalTok{]}
\DecValTok{7}         \FloatTok{0.92}      \FloatTok{0.90}       \FloatTok{0.02}\NormalTok{  [}\SpecialCharTok{{-}}\FloatTok{0.02}\NormalTok{, }\FloatTok{0.06}\NormalTok{]  }\FloatTok{1.02}\NormalTok{ [}\FloatTok{0.98}\NormalTok{, }\FloatTok{1.07}\NormalTok{]}
\DecValTok{8}         \FloatTok{0.09}      \FloatTok{0.08}       \FloatTok{0.01}\NormalTok{  [}\SpecialCharTok{{-}}\FloatTok{0.01}\NormalTok{, }\FloatTok{0.03}\NormalTok{]  }\FloatTok{1.12}\NormalTok{ [}\FloatTok{0.89}\NormalTok{, }\FloatTok{1.43}\NormalTok{]}
\DecValTok{9}         \FloatTok{0.87}      \FloatTok{0.88}      \SpecialCharTok{{-}}\FloatTok{0.01}\NormalTok{  [}\SpecialCharTok{{-}}\FloatTok{0.05}\NormalTok{, }\FloatTok{0.03}\NormalTok{]  }\FloatTok{0.99}\NormalTok{ [}\FloatTok{0.94}\NormalTok{, }\FloatTok{1.04}\NormalTok{]}
\DecValTok{10}        \FloatTok{1.03}      \FloatTok{3.24}      \SpecialCharTok{{-}}\FloatTok{2.21}\NormalTok{ [}\SpecialCharTok{{-}}\FloatTok{4.38}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.04}\NormalTok{]  }\FloatTok{0.32}\NormalTok{ [}\FloatTok{0.15}\NormalTok{, }\FloatTok{0.68}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

Users can also compute individual metrics using functions like
\texttt{eval\_eq\_opp()} to test specific fairness conditions. Full
usage examples are provided in the package documentation.

\section{Related Work}\label{related-work}

Other R packages similar to \texttt{fairmetrics} include
\texttt{fairness} (Kozodoi and V. Varga 2021), \texttt{fairmodels}
(Wiśniewski and Biecek 2022) and \texttt{mlr3fairness} (Pfisterer, Siyi,
and Lang 2024). \texttt{fairmetrics} differs from these packages in two
ways. The first difference is that \texttt{fairmetrics} calculates ratio
and difference-based group fairness metrics and their corresponding
confidence intervals, allowing for more meaningful inferences about the
fairness criteria. The second difference is that \texttt{fairmetrics}
does not possess any external dependencies and has a lower memory
footprint.
\hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows
the comparison of memory used and dependencies required when loading
each library.

\begin{table}[ht]
\centering
\begin{tabular}{l r r}
\hline
\textbf{Package} & \textbf{Memory (MB)} & \textbf{Dependencies} \\
\hline
fairmodels  & 17.02  & 29 \\
fairness    & 117.61 & 141\\
mlr3fairness & 58.11  & 45 \\
fairmetrics & 0.05   & 0  \\
\hline
\end{tabular}
\caption{Memory usage (in MB) and dependencies of `fairmetrics` vs similar packages.}
\label{tab:memory_dep_usage}
\end{table}

For Python users, the \texttt{fairlearn} library (Weerts et al. 2023)
provides additional fairness metrics and algorithms. The
\texttt{fairmetrics} package is designed for seamless integration with R
workflows, making it a more convenient choice for R users.

\section{Licensing and Availability}\label{licensing-and-availability}

The \texttt{fairmetrics} package is under the MIT license. It is
available on CRAN and can be installed by using
\texttt{install.packages("fairmetrics")}. Full documentation and its
examples are available at:
\url{https://jianhuig.github.io/fairmetrics/articles/fairmetrics.html}.
Source code and issue tracking are hosted on GitHub:
\url{https://github.com/jianhuig/fairmetrics/}.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024}
Gao, Jianhui, Benson Chou, Zachary R. McCaw, Hilary Thurston, Paul
Varghese, Chuan Hong, and Jessica Gronsbell. 2024. {``What Is Fair?
Defining Fairness in Machine Learning for Health.''} \emph{arXiv.org}.
\url{https://doi.org/10.48550/arXiv.2406.09307}.

\bibitem[\citeproctext]{ref-goldberger2000physiobank}
Goldberger, Ary L., Luis A. N. Amaral, Leon Glass, Jeffrey M. Hausdorff,
Plamen Ch. Ivanov, Roger G. Mark, Joseph E. Mietus, George B. Moody,
Chung-Kang Peng, and H. Eugene Stanley. 2000. {``PhysioBank,
PhysioToolkit, and PhysioNet: Components of a New Research Resource for
Complex Physiologic Signals.''} \emph{Circulation {[}Online{]}} 101
(23): e215--20. \url{https://doi.org/10.1161/01.CIR.101.23.e215}.

\bibitem[\citeproctext]{ref-fairness_package}
Kozodoi, Nikita, and Tibor V. Varga. 2021. \emph{Fairness: Algorithmic
Fairness Metrics}. \url{https://doi.org/10.32614/cran.package.fairness}.

\bibitem[\citeproctext]{ref-mattuMachineBias}
Mattu, Lauren Kirchner, Jeff Larson. 2016. {``Machine {Bias}.''}
\emph{ProPublica}.
https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing.

\bibitem[\citeproctext]{ref-mehrabi_survey_21}
Mehrabi, Ninareh, Fred Morstatter, Nripsuta Saxena, Kristina Lerman, and
Aram Galstyan. 2021. {``A Survey on Bias and Fairness in Machine
Learning.''} \emph{ACM Comput. Surv.} 54 (6).
\url{https://doi.org/10.1145/3457607}.

\bibitem[\citeproctext]{ref-mlr3fairness_package}
Pfisterer, Florian, Wei Siyi, and Michel Lang. 2024. \emph{Mlr3fairness:
Fairness Auditing and Debiasing for 'Mlr3'}.
\url{https://doi.org/10.32614/cran.package.mlr3fairness}.

\bibitem[\citeproctext]{ref-raffa2016clinical}
Raffa, Jesse. 2016. {``Clinical Data from the MIMIC-II Database for a
Case Study on Indwelling Arterial Catheters (Version 1.0).''}
\url{https://doi.org/10.13026/C2NC7F}.
\url{https://doi.org/10.13026/C2NC7F}.

\bibitem[\citeproctext]{ref-raffa2016data}
Raffa, Jesse D., Mohammad Ghassemi, Tristan Naumann, Mengling Feng, and
Daniel J. Hsu. 2016. {``Data Analysis.''} In \emph{Secondary Analysis of
Electronic Health Records}, 109--22. Springer, Cham.
\url{https://doi.org/10.1007/978-3-319-43742-2_9}.

\bibitem[\citeproctext]{ref-fairlearn_paper}
Weerts, Hilde, Miroslav Dudík, Richard Edgar, Adrin Jalali, Roman Lutz,
and Michael Madaio. 2023. {``FairLearn: Assessing and Improving Fairness
of AI Systems.''} \emph{arXiv.org}.
\url{https://doi.org/10.48550/arXiv.2303.16626}.

\bibitem[\citeproctext]{ref-wisniewski2022fairmodels}
Wiśniewski, Jakub, and Przemysław Biecek. 2022. {``Fairmodels: A
Flexible Tool for Bias Detection, Visualization, and Mitigation in
Binary Classification Models.''} \emph{The R Journal} 14 (1): 227--43.
\url{https://doi.org/10.32614/RJ-2022-019}.

\end{CSLReferences}

\end{document}

### 2025-08-21 R/get_fairness_metrics.R

@@ -29,11 +29,7 @@
 #' @param alpha Significance level for confidence intervals. Default is 0.05.
 #' @param digits Number of digits to round the metrics to. Default is 2.
 #'
-#' @return A list containing:
-#' \describe{
-#'   \item{performance}{Data frame with performance metrics by group.}
-#'   \item{fairness}{Data frame with computed fairness metrics and optional confidence intervals.}
-#' }
+#' @return A dataframe containing fairness assessments, the performance metrics they use and the evaluated results for each (binary) group (specified by the `group` parameter) along with the difference and ratio between them. If `confint` is  set to `TRUE`, then the estimated `(1-alpha)*100%` bootstrap confidence intervals are returned as well.
 #'
 #' @examples
 #' \donttest{
@@ -364,7 +360,20 @@ get_fairness_metrics <- function(data,
     lower_ratio_cis <-  round(exp(log(ratio_vals) - qnorm(1 - alpha / 2) * ratio_sds), digits)
     upper_ratio_cis <-  round(exp(log(ratio_vals) + qnorm(1 - alpha / 2) * ratio_sds), digits)
 
-    model_summary <- data.frame(
+
+    fairness_summary <- data.frame(
+      Fairness_Assesment = c(
+        "Statistical Parity",
+        "Equal Opportunity",
+        "Predictive Equality",
+        "Balance for Positive Class",
+        "Balance for Negative Class",
+        "Positive Predictive Parity",
+        "Negative Predictive Parity",
+        "Brier Score Parity",
+        "Overall Accuracy Parity",
+        "Treatment Equality"
+      ),
       Metric = c(
         "Positive Prediction Rate",
         "False Negative Rate",
@@ -381,12 +390,28 @@ get_fairness_metrics <- function(data,
         x[[1]]),
       Group2 = sapply(metrics, function(x)
         x[[2]]),
+      Difference = diff_vals,
+      Diff_CI  = paste0("[", lower_cis, ", ", upper_cis, "]"),
+      Ratio = round(ratio_vals, digits),
+      Ratio_CI = paste0("[", lower_ratio_cis, ", ", upper_ratio_cis, "]"),
       row.names = NULL
 
     )
 
+    colnames(fairness_summary) <- c(
+      "Fairness Assesment",
+      "Metric",
+      paste0("Group", sort(unique(data[[group]]))[1]),
+      paste0("Group", sort(unique(data[[group]]))[2]),
[…]

### 2025-08-22 R/get_fairness_metrics.R
typo fix

@@ -19,6 +19,8 @@
 #' - **Overall Accuracy Parity**: Difference in overall accuracy across groups.
 #' - **Treatment Equality**: Ratio of false negatives to false positives across groups.
 #'
+#' **NOTE:** Statistical inference from bootstrapped confidence intervals should be interpreted with caution. A confidence interval crossing 0 (for differences) or 1 (for ratios) means the evidence is inconclusive rather than proving absence of unfairness. Apparent violations may reflect sampling variability rather than systematic bias. Always complement these results with domain knowledge, sensitivity analyses, and additional fairness diagnostics before drawing strong conclusions about a specific fairness assesment.
+#'
 #' @param data A data frame containing the outcome, group, and predicted probabilities.
 #' @param outcome The name of the column containing the true binary outcome.
 #' @param group The name of the column representing the binary protected attribute (e.g., race, gender).

### 2025-08-22 R/get_fairness_metrics.R
another typo fix

@@ -19,7 +19,7 @@
 #' - **Overall Accuracy Parity**: Difference in overall accuracy across groups.
 #' - **Treatment Equality**: Ratio of false negatives to false positives across groups.
 #'
-#' **NOTE:** Statistical inference from bootstrapped confidence intervals should be interpreted with caution. A confidence interval crossing 0 (for differences) or 1 (for ratios) means the evidence is inconclusive rather than proving absence of unfairness. Apparent violations may reflect sampling variability rather than systematic bias. Always complement these results with domain knowledge, sensitivity analyses, and additional fairness diagnostics before drawing strong conclusions about a specific fairness assesment.
+#' **NOTE:** Statistical inference from bootstrapped confidence intervals should be interpreted with caution. A confidence interval crossing 0 (for differences) or 1 (for ratios) means the evidence is inconclusive rather than proving absence of unfairness. Apparent violations may reflect sampling variability rather than systematic bias. Always complement these results with domain knowledge, sensitivity analyses, and additional fairness diagnostics before drawing strong conclusions about a specific fairness assessment.
 #'
 #' @param data A data frame containing the outcome, group, and predicted probabilities.
 #' @param outcome The name of the column containing the true binary outcome.

### 2025-08-22 tests/testthat/helper-match-metrics.R
fixing tests

@@ -15,11 +15,11 @@ test_data <- mimic_preprocessed |>
 
 test_data$pred <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
 
-get_fairness_metrics(
+performance <- get_fairness_metrics(
   data = test_data,
   outcome = "day_28_flg",
   group = "gender",
   probs = "pred",
   cutoff = 0.41,
   confint = FALSE
-) |> list2env(envir = environment())
+)

### 2025-08-22 tests/testthat/test-Fairness-match-metrics.R

@@ -12,7 +12,7 @@ test_that('eval_acc_parity match metrics test', {
     rownames(res)<- 1
   },
   {
-    res <- subset(performance, Metric =="Accuracy")
+    res <- subset(performance, Metric =="Accuracy")[,2:4]
     rownames(res)<- 1
 
   })
@@ -32,7 +32,7 @@ test_that('eval_bs_parity match metrics test', {
     rownames(res)<- 1
   },
   {
-    res <- subset(performance, Metric =="Brier Score")
+    res <- subset(performance, Metric =="Brier Score")[,2:4]
     rownames(res)<- 1
   })
 })
@@ -52,7 +52,7 @@ test_that('eval_eq_opp match metrics test', {
     rownames(res)<- 1
     },
     {
-      res <- subset(performance, Metric =="False Negative Rate")
+      res <- subset(performance, Metric =="False Negative Rate")[,2:4]
       rownames(res)<- 1
     }
     )
@@ -75,7 +75,7 @@ test_that('eval_neg_class_bal match metrics test', {
     rownames(res) <- 1
   },
   {
-    res <- subset(performance, Metric == "Avg. Predicted Prob.")[2,]
+    res <- subset(performance, Metric == "Avg. Predicted Negative Prob.")[,2:4]
     rownames(res) <- 1
   }
   )
@@ -95,7 +95,7 @@ test_that('eval_pos_class_bal match metrics test', {
     rownames(res) <- 1
   },
   {
-    res <- subset(performance, Metric =="Avg. Predicted Prob.")[1,]
+    res <- subset(performance, Metric =="Avg. Predicted Positive Prob.")[,2:4]
     rownames(res) <- 1
   }
   )
@@ -116,7 +116,7 @@ test_that('eval_pred_equality metrics test', {
     rownames(res) <- 1
   },
   {
-    res <- subset(performance, Metric =="False Positive Rate")
+    res <- subset(performance, Metric =="False Positive Rate")[,2:4]
     rownames(res) <- 1
   }
   )
@@ -137,7 +137,7 @@ test_that('eval_stats_parity metrics test', {
     rownames(res) <- 1
   },
   {
-    res <- subset(performance, Metric =="Positive Prediction Rate")[1,]
+    res <- subset(performance, Metric =="Positive Prediction Rate")[,2:4]
     rownames(res) <- 1
   }
   )
@@ -159,7 +159,7 @@ test_that('eval_treatment_equality metrics test', {
[…]

### 2025-09-10 utils/joss_paper/paper.Rmd
fixing typo

@@
 affiliations:
-  - name: "Department of Statistical Science, University of Toronto"
+  - name: "Department of Statistical Sciences, University of Toronto"
     index: 1
@@
 output:
-  rticles::joss_article:
-    keep_tex: true
   md_document:
@@
      variant: "markdown_strict"
+  rticles::joss_article:
+    keep_tex: true
 journal: JOSS
@@
 
-           Fairness Assesment                                  Metric
+           Fairness Assessment                                 Metric
 1          Statistical Parity                Positive Prediction Rate

### 2025-09-10 R/get_fairness_metrics.R

@@ -401,7 +401,7 @@ get_fairness_metrics <- function(data,
     )
 
     colnames(fairness_summary) <- c(
-      "Fairness Assesment",
+      "Fairness Assessment",
       "Metric",
       paste0("Group", sort(unique(data[[group]]))[1]),
       paste0("Group", sort(unique(data[[group]]))[2]),

### 2025-09-11 utils/joss_paper/paper.tex
small fix 'Department of Statistical Sciences'
fixed references

@@
 
-      \affil[1]{Department of Statistical Science, University of
+      \affil[1]{Department of Statistical Sciences, University of
 Toronto}
@@
 Mart for Intensive Care, version II (MIMIC-II) database (Goldberger et
-al. 2000; J. Raffa 2016) to demonstrate its use.
+al. 2000; J. Raffa 2016; J. D. Raffa et al. 2016) to demonstrate its
+use.
 
@@
 
-\NormalTok{           Fairness Assesment                                  Metric}
+\NormalTok{           Fairness Assessment                                 Metric}
 \DecValTok{1}\NormalTok{          Statistical Parity                Positive Prediction Rate}
@@
 \bibitem[\citeproctext]{ref-fairness_package}
-Kozodoi, Nikita, and Tibor V. Varga. 2021. \emph{Fairness: Algorithmic
+Kozodoi, Nikita, and Tibor V. Varga. 2021. \emph{{fairness}: Algorithmic
 Fairness Metrics}. \url{https://doi.org/10.32614/cran.package.fairness}.
@@
 \emph{ProPublica}.
-https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing.
+\url{https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}.
 
@@
 \bibitem[\citeproctext]{ref-mlr3fairness_package}
-Pfisterer, Florian, Wei Siyi, and Michel Lang. 2024. \emph{Mlr3fairness:
-Fairness Auditing and Debiasing for 'Mlr3'}.
+Pfisterer, Florian, Wei Siyi, and Michel Lang. 2024.
+\emph{{mlr3fairness}: Fairness Auditing and Debiasing for 'Mlr3'}.
 \url{https://doi.org/10.32614/cran.package.mlr3fairness}.
@@
 Case Study on Indwelling Arterial Catheters (Version 1.0).''}
-\url{https://doi.org/10.13026/C2NC7F}.
 \url{https://doi.org/10.13026/C2NC7F}.
@@
 \bibitem[\citeproctext]{ref-wisniewski2022fairmodels}
-Wiśniewski, Jakub, and Przemysław Biecek. 2022. {``Fairmodels: A
+Wiśniewski, Jakub, and Przemysław Biecek. 2022. {``{fairmodels}: A
 Flexible Tool for Bias Detection, Visualization, and Mitigation in

### 2025-09-15 utils/joss_paper/paper.Rmd
using the abbreviation 'CI' for confidence interval

@@
 The package provides both point and interval estimates for a variety of commonly used criteria.
-`fairmetrics` also includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical] to demonstrate its use.
+`fairmetrics` also includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical; @raffa2016data] to demonstrate its use.
 
@@
 This limits users' ability to determine whether observed disparities are statistically significant.
-`fairmetrics` addresses this limitation by including confidence intervals for both difference and ratio based fairness metrics to enable more robust and statistically grounded fairness assessments.
+`fairmetrics` addresses this limitation by including confidence intervals (CIs) for both difference and ratio based fairness metrics to enable more robust and statistically grounded fairness assessments.
 
@@
 The model is trained on a subset of the data and the predictions are made and evaluated on a test set.
-A statistically significant difference across groups at a given level of significance is indicated when the confidence interval for a difference-based metric does not include zero or when the interval for a ratio-based metric does not include one.
+A statistically significant difference across groups at a given level of significance is indicated when the CI for a difference-based metric does not include zero or when the interval for a ratio-based metric does not include one.
 
@@
 library(fairmetrics)
-# Setting alpha=0.05 for 95% confidence intervals
+# Setting alpha=0.05 for 95% CIs
 get_fairness_metrics(
@@
 `fairmetrics` differs from these packages in two ways.
-The first difference is that `fairmetrics` calculates ratio and difference-based group fairness metrics and their corresponding confidence intervals, allowing for more meaningful inferences about the fairness criteria.
+The first difference is that `fairmetrics` calculates ratio and difference-based group fairness metrics and their corresponding CIs, allowing for more meaningful inferences about the fairness criteria.
 The second difference is that `fairmetrics` does not possess any external dependencies and has a lower memory footprint.

### 2025-09-15 utils/joss_paper/paper.tex
recommitting the markdown text

@@
 observed disparities are statistically significant. \texttt{fairmetrics}
-addresses this limitation by including confidence intervals for both
-difference and ratio based fairness metrics to enable more robust and
-statistically grounded fairness assessments.
+addresses this limitation by including confidence intervals (CIs) for
+both difference and ratio based fairness metrics to enable more robust
+and statistically grounded fairness assessments.
 
@@
 significant difference across groups at a given level of significance is
-indicated when the confidence interval for a difference-based metric
-does not include zero or when the interval for a ratio-based metric does
-not include one.
+indicated when the CI for a difference-based metric does not include
+zero or when the interval for a ratio-based metric does not include one.
 
@@
 \FunctionTok{library}\NormalTok{(fairmetrics)}
-\CommentTok{\# Setting alpha=0.05 for 95\% confidence intervals}
+\CommentTok{\# Setting alpha=0.05 for 95\% CIs}
 \FunctionTok{get\_fairness\_metrics}\NormalTok{(}
@@
 ways. The first difference is that \texttt{fairmetrics} calculates ratio
-and difference-based group fairness metrics and their corresponding
-confidence intervals, allowing for more meaningful inferences about the
-fairness criteria. The second difference is that \texttt{fairmetrics}
-does not possess any external dependencies and has a lower memory
-footprint.
+and difference-based group fairness metrics and their corresponding CIs,
+allowing for more meaningful inferences about the fairness criteria. The
+second difference is that \texttt{fairmetrics} does not possess any
+external dependencies and has a lower memory footprint.
 \hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows

### 2025-09-15 utils/joss_paper/paper.Rmd
removing comments in code

@@
 output:
+  rticles::joss_article:
+    keep_tex: true
   md_document:
@@
      variant: "markdown_strict"
-  rticles::joss_article:
-    keep_tex: true
 journal: JOSS
@@
 ```r
-# Train a classification model (e.g., random forest).
-# Add the vector of predicted probabilities to the test data
-# to evaluate fairness.
 library(fairmetrics)
